# Where and when the timetable data actually exists

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.path = "figures/coverage-", dpi = 110,
                      fig.width = 9, fig.height = 5)
library(ggplot2)
library(data.table)
library(sf)

cv <- readRDS(coverage_path)
areas <- coverage_atco_areas()

fmt <- function(x) format(round(x), big.mark = ",", scientific = FALSE,
                          trim = TRUE)
kt <- function(x, ...) knitr::kable(x, format = "markdown", ...)

STATUS <- c("present", "partial", "severe shortfall", "absent",
            "not yet coded separately", "no archive")
PAL <- c(present = "#cfe8cf", partial = "#ffe2a8",
         `severe shortfall` = "#f7b799", absent = "#f2a0a0",
         `not yet coded separately` = "#cdd7ea",
         `no archive` = "#d9d9d9")

cov <- copy(cv$coverage)
reg <- copy(cv$region)
nat <- copy(cv$national)
gaps <- copy(cv$gaps)
gap_years <- sort(unique(gaps$year))
```

The manual's coverage figure — `images/manual/transport_bus_data.webp` — is a
hand-assessed grid of region against year, green for usable and red for not.
It was right about the big things and it is the reason nobody reads 2004
Scotland as a collapse in bus service. But it was compiled by eye several
years ago, it stops at 2023, and it records eleven regions, which is coarse
enough to hide an entire county.

This report rebuilds it by measurement, extends it to `r max(cv$years)`, and
puts it in the pipeline so it stays current. It is generated from the
converted feeds themselves, so a reconversion re-measures it rather than
leaving the assessment to drift away from the data.

## How coverage is measured

The unit is the **ATCO administrative area**, not the region. A coverage gap
is physically a set of files that is not in the archive, and the archives are
assembled per administrative area, so that is the unit at which data goes
missing. Every stop id in every era carries its area in the first three
characters of the NaPTAN AtcoCode, and between 99.6% and 100% of stops in
every feed join to UK2GTFS's `atco_areas` lookup, so one rule works across
NPTDR, the Bus Archive and TNDS alike. Boundaries come from that same lookup.

Three numbers are taken per area per year, because they fail differently:
how many stops the feed **describes**, how many of those a vehicle actually
**calls at**, and how many **departures** that amounts to. A feed can list an
area's stops and run nothing from them, which a stop count alone scores as
present. Departures is the measure used throughout.

An area is judged against **its own usual level**, and on its **share of that
year's departures** rather than the raw count. The eras are not on a common
scale: a Bus Archive year merges four weekly snapshots, so 2014–2017 hold
around 220 million departures against 28–71 million in an NPTDR or TNDS year.
Any benchmark in raw departures inherits that multiplier. A share cancels it,
and it also cancels genuine national change — which is what is wanted here,
because this measures coverage, not service. `absent` is no departures at
all, `severe shortfall` is below a quarter of the area's usual share, and
`partial` is below six tenths.

**An area missing because its stops are filed under a predecessor authority
is not a coverage gap**, and is shown separately in blue. Local government
reorganisation creates ATCO areas mid-series, and the service was in the
archive all along under the old code. The test is whether the *region* grew
when the area first appeared: a gap filling in adds the area's stops to its
region, a re-coding only moves them within it.

Only the **bus-side feeds** are measured. The rail CIF feeds are keyed on
TIPLOC rather than ATCO, so the same join cannot be made; rail coverage is a
separate question with a separate source.

One caution remains. The method cannot tell a missing file from a real
collapse in service, so a `partial` year may be a genuine reduction — 2020 is
the obvious case. Only `absent` is unambiguous.

## What each year is made of

```{r sources}
f <- copy(cv$feeds)
f[, src := fcase(grepl("nptdr", path), "NPTDR",
                 grepl("busarchive", path), "Bus Archive (TNDS weekly)",
                 grepl("bods_coach", path), "BODS Coach",
                 grepl("tnds", path), "TNDS",
                 default = "other")]
tab <- f[, list(source = paste(unique(src), collapse = " + "),
                window = paste0(min(window_start), " to ", max(window_end))),
         by = "year"][order(year)]
kt(tab, col.names = c("year", "source", "28-day counting window"))
```

`r if (length(cv$missing_years)) paste0("There is no archive at all for ", paste(cv$missing_years, collapse = " and "), ": NPTDR stopped after 2011 and the Bus Archive starts in 2014. Those years are shown as a grey column throughout and are not interpolated anywhere in this pipeline.") else ""`

## The coverage grid

This is the direct successor to the manual's figure: region by year, coloured
by measured status rather than by assessment.

```{r grid, fig.height=4.2}
allyr <- min(cv$years):max(cv$years)
g <- CJ(year = allyr, region = sort(unique(reg$region)), unique = TRUE)
g <- merge(g, reg[, list(year, region, status)], by = c("year", "region"),
           all.x = TRUE)
g[is.na(status), status := "no archive"]
g[, status := factor(status, levels = STATUS)]
g[, region := factor(region, levels = rev(sort(unique(region))))]

ggplot(g, aes(factor(year), region, fill = status)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  scale_fill_manual(values = PAL, drop = FALSE, name = NULL) +
  labs(x = NULL, y = NULL,
       title = "Bus timetable coverage by region and year",
       subtitle = "measured from the converted feeds; departures against each region's own median") +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "bottom")
```

The region roll-up is generous by construction: a region keeps most of its
departures when one of its areas vanishes, so a region only turns amber when
a large area or several small ones go. The area-level maps below are what the
regional view hides.

### How this compares with the manual's figure

```{r london, results='asis'}
ld <- reg[region == "London"][order(year)]
bad <- ld[status == "severe shortfall"]
cat("The strongest agreement is London. The manual marks it red in ",
    "2004-2006 and 2014-2017, and those are exactly the seven years this ",
    "finds: London runs at ", round(100 * min(bad$index)), "-",
    round(100 * max(bad$index)), "% of its usual share of national ",
    "departures in them, against 91-102% in 2007-2011 and 112-143% from ",
    "2018. Whatever else is true of those years, the London bus network is ",
    "not in the archive in any useful quantity.\n", sep = "")
```

The two do not agree everywhere, and they are not measuring the same thing.
The manual records a judgement about whether a region's data is *usable*,
which takes in problems this cannot see — a snapshot taken at the wrong time
of year, a known conversion failure, an operator's data arriving in the wrong
format. This measures one narrow, checkable thing: whether departures are
present in the quantity the area normally has.

So a region-year the manual flags and this one does not has probably not been
disproved; it has been found to contain data, which was never the whole of
the manual's claim. The reverse case — flagged here and not there — is the
one worth acting on, because it means something measurable changed since the
manual was written, or was missed by eye.

## Coverage in space

```{r maps, fig.height=11, fig.width=9}
# as.data.frame on the right-hand side: merging an sf with a data.table can
# come back without the sf class, and then geom_sf has nothing to draw
m <- areas[areas$region != "Great Britain", ]
md <- merge(m, as.data.frame(cov[, list(atco_code, year, status)]),
            by = "atco_code")
md$status <- factor(md$status, levels = STATUS)

ggplot(md) +
  geom_sf(aes(fill = status), colour = "grey40", linewidth = 0.05) +
  facet_wrap(~ year, ncol = 5) +
  scale_fill_manual(values = PAL, drop = FALSE, name = NULL) +
  labs(title = "Bus timetable coverage by ATCO administrative area",
       subtitle = "one panel per analysis year; 2012 and 2013 have no archive") +
  theme_void(base_size = 9) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 9, face = "bold"))
```

## The years with real gaps

```{r gapsummary, results='asis'}
if (nrow(gaps) == 0) {
  cat("No area falls below a quarter of its usual service in any year.\n")
} else {
  s <- gaps[, list(areas = uniqueN(atco_code),
                   absent = uniqueN(atco_code[status == "absent"]),
                   regions = paste(sort(unique(region)), collapse = ", ")),
            by = "year"][order(year)]
  print(kt(s, col.names = c("year", "areas affected", "of which absent",
                            "regions")))
}
```

```{r worstyear, results='asis'}
wy <- cv$national[order(areas_absent, decreasing = TRUE)][1]
cat("\n**", wy$year, "is the worst year**: ", wy$areas_absent,
    " areas have no service at all, against ", wy$areas_present,
    " present.\n", sep = "")
```

### `r cv$national[order(-areas_absent)][1]$year` in detail

```{r worstdetail}
wyr <- cv$national[order(-areas_absent)][1]$year
d <- cov[year == wyr & status == "absent"][order(region, name)]
kt(d[, list(region, name, atco_code,
            typical_departures = fmt(typical))],
   col.names = c("region", "area", "ATCO", "departures in a normal year"))
```

```{r worstmap, fig.height=6, fig.width=6.5}
one <- merge(m, as.data.frame(cov[year == wyr, list(atco_code, status)]),
             by = "atco_code")
one$status <- factor(one$status, levels = STATUS)
ggplot(one) +
  geom_sf(aes(fill = status), colour = "grey40", linewidth = 0.1) +
  scale_fill_manual(values = PAL, drop = FALSE, name = NULL) +
  labs(title = paste0("Coverage in ", wyr),
       subtitle = "red areas contribute no bus service to that year's outputs") +
  theme_void(base_size = 11) +
  theme(legend.position = "bottom")
```

### Absences that are not gaps

```{r recoded, results='asis'}
rc <- cv$recoded
if (is.null(rc) || nrow(rc) == 0) {
  cat("No area's absence is explained by a later re-coding.\n")
} else {
  cat("These areas are missing from the early years because they did not ",
      "exist yet as ATCO areas. Their stops are in the archive under the ",
      "predecessor authority, and the region does not grow when they ",
      "appear — so they are **not** counted as gaps anywhere above.\n\n",
      sep = "")
  print(kt(rc[order(appears, name)][, list(region, name, atco_code,
                                           years, appears)],
           col.names = c("region", "area", "ATCO", "absent from",
                         "first appears")))
}
```

## Coverage in time

```{r natseries, fig.height=4}
n <- melt(nat[, list(year, `areas present` = areas_present,
                     `not yet coded separately` = areas_recoded,
                     `areas absent` = areas_absent)],
          id.vars = "year", variable.name = "k", value.name = "v")
ggplot(n, aes(year, v, fill = k)) +
  geom_col() +
  scale_fill_manual(values = c(`areas present` = "#cfe8cf",
                               `not yet coded separately` = "#cdd7ea",
                               `areas absent` = "#f2a0a0"), name = NULL) +
  scale_x_continuous(breaks = cv$years) +
  labs(x = NULL, y = "ATCO areas",
       title = "How many administrative areas each year's archive contains") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "bottom", panel.grid.minor = element_blank())
```

```{r depseries, fig.height=4.6}
ggplot(nat, aes(year, departures / 1e6)) +
  geom_line(linewidth = 0.8, colour = "grey30") +
  geom_point(size = 1.8, colour = "grey20") +
  scale_x_continuous(breaks = cv$years) +
  labs(x = NULL, y = "million departures in the feed",
       title = "Total scheduled departures in each year's bus feed",
       subtitle = "the feed as converted, before the 28-day window and mode filters") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        panel.grid.minor = element_blank())
```

A change in this line is not by itself a change in bus service: it moves with
the source, with the archive's completeness, and with how much of the network
a snapshot happened to catch. It is here to be read against the grid above,
not on its own.

```{r regseries, fig.height=6.5}
r <- reg[region != "Unassigned"]
ggplot(r, aes(year, departures / 1e6)) +
  geom_line(linewidth = 0.6, colour = "grey30") +
  geom_point(data = r[status %in% c("absent", "severe shortfall")],
             colour = "#c0392b", size = 1.6) +
  facet_wrap(~ region, scales = "free_y", ncol = 3) +
  scale_x_continuous(breaks = seq(2005, 2025, by = 5)) +
  labs(x = NULL, y = "million departures",
       title = "Departures by region",
       subtitle = "red points mark a region below a quarter of its usual level") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

## Every gap, by area and year

```{r gaptable}
if (nrow(gaps) > 0) {
  t2 <- gaps[, list(year, region, area = name, atco = atco_code,
                    departures = fmt(departures),
                    normal = fmt(typical),
                    pct = paste0(round(100 * index), "%"),
                    status)][order(year, region, area)]
  kt(t2, col.names = c("year", "region", "area", "ATCO", "departures",
                       "normal", "share of normal", "status"))
}
```

## How to read the series because of this

The gaps are not noise spread evenly over the country. They are whole
administrative areas, so they land as a contiguous block on a map, and any
statistic that sums or averages over zones will move when one appears or
disappears. Three consequences:

**A year with a gap is not comparable with a year without one**, for any
national total, in any mode. This is most acute in the earliest year of the
series, which is also the one most often wanted as a baseline.

**Per-zone figures are safe where the zone has data.** A gap removes zones
from the series rather than depressing the ones that remain, so a like-for-
like comparison restricted to zones present in both years is unaffected. The
published change columns already avoid the worst of it by taking
`maxbus_2006_2008` rather than a single early year.

**A gap and a service cut look identical in a zone-count.** The only reliable
separator is the one used here: an entire administrative area going to zero
and returning is an archive artefact, because real networks do not do that.

## Prefixes that are not an administrative area

```{r unassigned, results='asis'}
ua <- cv$unassigned
if (is.null(ua) || nrow(ua) == 0) {
  cat("Every stop id prefix in every feed maps to a known ATCO area.\n")
} else {
  cat("A stop id whose first three characters are not a known ATCO area code.",
      " These are excluded from every count above: there are ", nrow(ua),
      " of them, they appear in one or two years each, and left in they would",
      " contribute a permanent baseline of absent \"areas\" that means",
      " nothing. Largest first.\n\n", sep = "")
  print(kt(head(ua[order(-stops)], 20),
           col.names = c("prefix", "years present", "most stops in a year",
                         "most departures")))
}
```

```{r footer, results='asis'}
cat("\n---\n\nGenerated ", format(cv$generated, "%Y-%m-%d %H:%M"),
    " from ", nrow(cv$feeds), " bus feeds covering ",
    length(cv$years), " years.\n", sep = "")
```
