# Where TNDS and the DfT's BODS GTFS disagree most, zone by zone

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.path = "figures/lsoagap-", dpi = 110,
                      fig.width = 9, fig.height = 5)
library(ggplot2)
library(data.table)

g <- readRDS(gap_path)
z <- as.data.table(g$zones)
top <- as.data.table(g$top)
svc <- as.data.table(g$services)
nat <- g$national

# scientific = FALSE: format() renders a round number such as 9000000 as 9e+06
# whatever big.mark says
fmt <- function(x) format(round(x), big.mark = ",", scientific = FALSE,
                          trim = TRUE)
pct <- function(x) sprintf("%.1f%%", 100 * x)
cn <- c(E = "England", W = "Wales", S = "Scotland")

theme_report <- theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "top",
        plot.title.position = "plot")
gap_cols <- c("#2a78d6", "#d94f70")

# Reused for both directions of the comparison: the zone table and, under it,
# the services that make up each zone's gap.
zone_table <- function(d) {
  data.frame(
    Zone = d$zone_id,
    Locality = ifelse(is.na(d$locality), "—", d$locality),
    Country = cn[d$country],
    TNDS = round(d$runs_tnds),
    `BODS GTFS` = round(d$runs_bods),
    Difference = round(d$gap),
    `Only TNDS` = round(d$gap_services_only_tnds),
    `Only BODS` = round(d$gap_services_only_bods),
    Frequency = round(d$gap_frequency),
    check.names = FALSE)
}

service_table <- function(zid, n = 8) {
  d <- svc[zone_id == zid]
  setorderv(d, "gap", -1L)
  d <- d[order(-abs(gap))][seq_len(min(n, .N))]
  data.frame(
    Service = ifelse(is.na(d$name) | d$name == "", "(unnamed)", d$name),
    Description = substr(ifelse(is.na(d$long_name), "—", d$long_name), 1, 46),
    TNDS = round(d$tnds),
    `BODS GTFS` = round(d$bods_gtfs),
    Difference = round(d$gap),
    Cause = d$cause,
    check.names = FALSE)
}
```

The comparison report (`bus_source_comparison.md`) measures how far the three
bus timetable sources disagree nationally. This report asks a narrower and more
practical question: **for which individual LSOAs (Data Zones in Scotland) does
the choice between TNDS and the DfT's BODS GTFS change the answer most, and
which bus routes are responsible?**

Both sources are the **`r g$year`** snapshot, deduplicated with
`UK2GTFS::gtfs_deduplicate()` so that neither describes the same bus twice,
and counted over the same 28-day window (**`r g$window$startdate` to
`r g$window$enddate`**) on the **plain LSOA21 / DZ22 boundaries**. The
published trips-per-zone outputs use zones widened for stop access; this
report does not, because widened zones overlap and a disagreement at a stop
in several of them would be counted several times over:

- TNDS (TransXChange), converted with `UK2GTFS::transxchange2gtfs()`:
  `r basename(g$snapshot$tnds)`
- BODS GTFS, the DfT's own rendering, used as supplied:
  `r sub(".*GTFS/([0-9]+)/.*", "\\1", g$snapshot$bods_gtfs)`

The measure is **total bus trip-runs in the window**: every vehicle journey
that calls at least once inside the zone, weighted by the number of times it
runs over the 28 days. A journey counts once per zone however many of the
zone's stops it calls at, which is exactly what `gtfs_trips_per_zone()` does, so
these numbers reconcile with the comparison report's zone totals. Unlike
`tph_daytime_avg` — the measure the pipeline publishes — it includes the night
band and does not weight weekdays, so it is a plain total.

## The national picture

```{r national, results="asis"}
tot <- data.frame(
  Measure = c("Zones with counted bus service in either source",
              "Total bus trip-runs, TNDS",
              "Total bus trip-runs, BODS GTFS",
              "Zones where TNDS counts more",
              "Zones where BODS GTFS counts more",
              "Zones where the two agree exactly",
              "Zones with service in TNDS only",
              "Zones with service in BODS GTFS only"),
  Value = c(fmt(nat$zones), fmt(nat$runs_tnds), fmt(nat$runs_bods),
            fmt(nat$zones_tnds_higher), fmt(nat$zones_bods_higher),
            fmt(nat$zones_equal), fmt(nat$zones_only_tnds),
            fmt(nat$zones_only_bods)))
knitr::kable(tot, row.names = FALSE, align = c("l", "r"))
```

```{r national-note, results="asis"}
# Deliberately does not characterise the size of the national difference in
# words - an earlier draft called it "near-agreement", which was written before
# the number was known and was wrong by an order of magnitude.
higher <- if (nat$runs_tnds > nat$runs_bods) "TNDS" else "BODS GTFS"
cat(sprintf(paste0(
  "Nationally TNDS counts %s bus trip-runs against BODS GTFS's %s, so %s is ",
  "the higher of the two by %s (%+.1f%% of the BODS GTFS total). The direction ",
  "is not uniform: BODS GTFS is the higher source in %s zones and TNDS in %s, ",
  "so the national total is a partial cancellation of disagreements pointing ",
  "opposite ways.\n\n"),
  fmt(nat$runs_tnds), fmt(nat$runs_bods), higher,
  fmt(abs(nat$runs_tnds - nat$runs_bods)),
  100 * (nat$runs_tnds / nat$runs_bods - 1),
  fmt(nat$zones_bods_higher), fmt(nat$zones_tnds_higher)))

cat(sprintf(paste0(
  "At zone level the two agree exactly in only %s of %s zones (%s), and in ",
  "**%s zones** one source shows a bus service where the other shows none at ",
  "all — %s in TNDS only, %s in BODS GTFS only. Those are the zones where the ",
  "choice of source is not a matter of degree.\n"),
  fmt(nat$zones_equal), fmt(nat$zones), pct(nat$zones_equal / nat$zones),
  fmt(nat$zones_only_tnds + nat$zones_only_bods),
  fmt(nat$zones_only_tnds), fmt(nat$zones_only_bods)))
```

```{r gap-dist, fig.height=4.2}
zz <- copy(z)
zz[, band := cut(gap, breaks = c(-Inf, -2000, -500, -100, -1, 1, 100, 500,
                                 2000, Inf),
                 labels = c("< -2,000", "-2,000 to -500", "-500 to -100",
                            "-100 to -1", "equal", "1 to 100", "100 to 500",
                            "500 to 2,000", "> 2,000"))]
bd <- zz[, .N, by = band][order(band)]
ggplot(bd, aes(x = band, y = N,
               fill = ifelse(band == "equal", "equal",
                             ifelse(as.integer(band) < 5, "BODS GTFS higher",
                                    "TNDS higher")))) +
  geom_col(width = 0.72) +
  scale_fill_manual(values = c("BODS GTFS higher" = gap_cols[2],
                               "TNDS higher" = gap_cols[1],
                               "equal" = "grey65"), name = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(title = "How far the two sources differ, zone by zone",
       subtitle = "TNDS minus BODS GTFS bus trip-runs in the 28-day window",
       x = NULL, y = "Zones") +
  theme_report +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### Is the disagreement spread across the country or concentrated?

Every zone, ordered by its difference and plotted in that order: the far left is
where TNDS counts most more, the far right where the DfT's GTFS does. The
histogram above shows how many zones fall in each band; this shows the shape of
the whole distribution at once, and in particular whether the national totals
are driven by a long broad disagreement or by a short extreme tail.

```{r gap-curve, fig.height=4.6}
zc <- z[order(-gap)]
zc[, rank := seq_len(.N)]
zc[, side := ifelse(gap > 0, "TNDS higher",
                    ifelse(gap < 0, "BODS GTFS higher", "equal"))]
side_cols <- c("TNDS higher" = gap_cols[1], "BODS GTFS higher" = gap_cols[2],
               "equal" = "grey75")
comma <- function(x) format(x, big.mark = ",", scientific = FALSE, trim = TRUE)

ggplot(zc, aes(x = rank, y = gap, fill = side)) +
  geom_col(width = 1) +
  geom_hline(yintercept = 0, colour = "grey25", linewidth = 0.4) +
  scale_fill_manual(values = side_cols, name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0.005, 0.005)),
                     labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Every zone, ordered by how far the two sources disagree",
       subtitle = paste0("TNDS minus BODS GTFS bus trip-runs in the 28-day ",
                         "window; above the line TNDS counts more"),
       x = paste0("Zones, ordered by difference (", comma(nrow(zc)), " in all)"),
       y = "TNDS − BODS GTFS") +
  theme_report
```

On a linear axis the middle of that curve is flat whether the zones there agree
exactly or differ by a hundred departures, because the ends are thousands of
times larger. The same curve on a signed logarithmic axis separates the two —
each gridline is ten times the last, in both directions from zero:

```{r gap-curve-log, fig.height=4.6}
# Signed pseudo-log: sign(x) * log10(1 + |x|). Plain log cannot show negatives
# and cannot show zero, and both matter here.
slog <- function(x) sign(x) * log10(1 + abs(x))
brk <- c(-100000, -10000, -1000, -100, -10, 0, 10, 100, 1000, 10000, 100000)
brk <- brk[abs(brk) <= max(abs(zc$gap)) * 10]

ggplot(zc, aes(x = rank, y = slog(gap), fill = side)) +
  geom_col(width = 1) +
  geom_hline(yintercept = 0, colour = "grey25", linewidth = 0.4) +
  scale_fill_manual(values = side_cols, name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0.005, 0.005)),
                     labels = comma) +
  scale_y_continuous(breaks = slog(brk), labels = comma(brk)) +
  labs(title = "The same curve on a signed logarithmic scale",
       subtitle = paste0("Each gridline is ten times the one before; the flat ",
                         "middle of the linear plot opens up here"),
       x = paste0("Zones, ordered by difference (", comma(nrow(zc)), " in all)"),
       y = "TNDS − BODS GTFS") +
  theme_report
```

```{r gap-curve-text, results = 'asis'}
tot <- sum(abs(zc$gap))
share <- function(frac) {
  n <- ceiling(frac * nrow(zc))
  o <- sort(abs(zc$gap), decreasing = TRUE)
  sum(o[seq_len(n)]) / tot
}
band <- function(lo, hi) sum(abs(zc$gap) >= lo & abs(zc$gap) < hi)
# State the two measurements and let them sit next to each other. An earlier
# draft opened with "the disagreement is concentrated", written before the
# numbers were computed; the tail is heavy AND the middle is far from zero, and
# only one of those was in the sentence.
cat(sprintf(paste0(
  "Summed over every zone, the two sources differ by **%s trip-runs** in ",
  "absolute terms, against **%s** between the national totals — the ",
  "difference between those two figures is disagreement that cancels between ",
  "zones pointing opposite ways. Of the absolute total, the worst **1%%** of ",
  "zones carry **%s** and the worst **10%%** carry **%s**.\n\n"),
  fmt(tot), fmt(abs(nat$runs_tnds - nat$runs_bods)),
  pct(share(0.01)), pct(share(0.10))))
cat(sprintf(paste0(
  "By size of difference: **%s zones** differ by 1,000 trip-runs or more (%s ",
  "of all zones), **%s** by between 100 and 1,000, **%s** by between 1 and 100, ",
  "and **%s** agree exactly. A zone differing by 100 trip-runs over 28 days is ",
  "under four departures a day; one differing by 1,000 is thirty-six.\n\n"),
  fmt(band(1000, Inf)), pct(band(1000, Inf) / nrow(zc)),
  fmt(band(100, 1000)), fmt(band(1, 100)), fmt(sum(zc$gap == 0))))
cross <- which(zc$gap <= 0)[1]
cat(sprintf(paste0(
  "The curve is not symmetric. It crosses zero at rank **%s of %s**, %s of the ",
  "way along, and the two ends are of very different size: the highest zone is ",
  "**+%s** and the lowest **%s**, so the drop on the right is about %.0f times ",
  "the rise on the left. BODS GTFS counts more service than TNDS in more zones ",
  "and by a wider margin.\n\n"),
  fmt(cross), fmt(nrow(zc)), pct(cross / nrow(zc)),
  fmt(max(zc$gap)), fmt(min(zc$gap)), abs(min(zc$gap)) / max(zc$gap)))

# Counting zones and ranking them by size give different answers about where
# the TNDS-high side is, so report both rather than either alone.
cty <- function(d) {
  n <- d[, .N, by = country][order(-N)]
  paste(sprintf("%s %s", fmt(n$N), ifelse(is.na(cn[n$country]), "unknown",
                                          cn[n$country])), collapse = ", ")
}
pos <- zc[gap > 0]
worst <- pos[seq_len(max(1, ceiling(0.01 * .N)))]
lead <- function(d) unname(cn[d[, .N, by = country][order(-N)][1]$country])
# Which country leads by count and which by size are separate questions, and on
# some snapshots they have the same answer - the sentence has to work either way
# rather than assume they differ.
verdict <- if (identical(lead(pos), lead(worst))) {
  sprintf(paste0("%s leads on both counts here, though that is a property of ",
                 "this snapshot rather than of the sources: on the February ",
                 "2026 feeds the extremes were mostly Scottish"), lead(pos))
} else {
  sprintf(paste0("%s has the most such zones and %s the most extreme ones, so ",
                 "counting zones and ranking them by size give different ",
                 "answers"), lead(pos), lead(worst))
}
cat(sprintf(paste0(
  "The left-hand side is not one country's story either way. Of the **%s ",
  "zones** where TNDS counts more, the split by country is %s; but among the ",
  "worst **1%%** of that side it is %s. %s. The top-ten tables below rank by ",
  "size, so they show only the second of those two answers.\n\n"),
  fmt(nrow(pos)), cty(pos), cty(worst), verdict))
cat(sprintf(paste0(
  "So the answer to \"a few extreme zones or many moderate ones\" is both, and ",
  "the two facts have to be held together: the tail is heavy enough that a ",
  "tenth of zones account for %s of all disagreement, yet **%s of zones** ",
  "differ by more than thirty-six departures a day, which is not a rounding ",
  "error in any of them. The choice of source changes the answer over most of ",
  "the country, and changes it drastically in a small part of it.\n\n"),
  pct(share(0.10)), pct(band(1000, Inf) / nrow(zc))))
```

## The zones that disagree most

Ranked on the absolute difference in trip-runs, both directions. The last three
columns split each zone's difference into services **only** one source carries
and services **both** carry at different frequencies — the same decomposition
the comparison report applies nationally, but computed within the zone. They are
signed contributions and add up to `Difference`, so "Only BODS" is negative:
service the DfT feed has and TNDS does not pulls the difference down.

Expect repeated localities. City-centre LSOAs are small and the zone polygons
are widened by 500 m, so half a dozen adjacent zones can all contain the same
bus station and all rank together on the same handful of services. That is a
faithful answer to "which zones disagree most" rather than a fault in the
ranking, but it means the tables describe fewer distinct places than rows. The
route-level section below therefore takes one zone per locality.

### Zones where TNDS counts substantially more service

```{r top-tnds, results="asis"}
t1 <- head(top[order(-gap)], 15)
knitr::kable(zone_table(t1), row.names = FALSE,
             format.args = list(big.mark = ","))
```

### Zones where BODS GTFS counts substantially more service

```{r top-bods, results="asis"}
t2 <- head(top[order(gap)], 15)
knitr::kable(zone_table(t2), row.names = FALSE,
             format.args = list(big.mark = ","))
```

```{r split-summary, results="asis"}
sp <- top[, list(
  only_tnds = sum(gap_services_only_tnds),
  only_bods = sum(gap_services_only_bods),
  freq = sum(gap_frequency))]
cat(sprintf(paste0(
  "Across the %d zones investigated in detail, the differences come to %s ",
  "trip-runs on services only TNDS carries, %s on services only BODS GTFS ",
  "carries, and a net %s from services both carry at different frequencies. ",
  "Missing services, not frequency differences, dominate.\n"),
  nrow(top), fmt(sp$only_tnds), fmt(-sp$only_bods), fmt(sp$freq)))
```

## Is either source still counting the same bus twice?

A zone's total can be inflated without any extra service existing, if the feed
publishes one journey more than once. Both feeds have already been through
`UK2GTFS::gtfs_deduplicate()`, so what follows measures what that deliberately
left behind, not the sources as published.

The test used here is the looser of the two: within each zone a journey is
identified by the (stop, departure time) pairs it makes at that zone's stops,
and two trips with the same route number, the same signature and the same
operating **date** are the same bus counted twice. Removal is stricter than
that — it needs the whole itinerary to match, the route and trip attributes to
agree, and every date of the copy removed to be covered by the copy kept.
Anything reported below therefore falls in the gap between the two: copies that
overlap in the window only partly, or that one source publishes under a
different route number or operator.

The date matters to both. GTFS models a school-term journey and its holiday
twin as two trips with identical times and complementary calendars, which is
correct modelling; a test that ignored dates would call every one of those a
duplicate.

```{r duplication-national, results="asis"}
dn <- as.data.table(g$duplication_national)
if (nrow(dn)) {
  src_lab <- c(tnds = "TNDS (TransXChange)", bods_gtfs = "BODS (GTFS)")
  # print(): a kable inside an if block is not the chunk's last expression, so
  # its value is discarded and the table silently never appears
  print(knitr::kable(
    data.frame(Source = src_lab[dn$source],
               `Bus trips` = fmt(dn$bus_trips),
               `Distinct journeys` = fmt(dn$distinct_journeys),
               `Trip-days` = fmt(dn$trip_days),
               `Duplicate runs` = fmt(dn$duplicate_runs),
               Share = pct(dn$share_duplicate),
               check.names = FALSE),
    row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
    caption = "Whole-feed duplicate journeys remaining, by source"))
  worse <- dn$source[which.max(dn$share_duplicate)]
  cat(sprintf(paste0(
    "\nNationally, **%s** of the counted runs still left in %s are the same ",
    "journey twice on one day, against %s in the other source. Across the ",
    "whole feed a journey is its entire itinerary, which is a stricter test ",
    "than the zone-level one below: inside a zone only the part of the trip ",
    "that touches the zone's stops can be compared.\n\n"),
    pct(max(dn$share_duplicate)), src_lab[worse],
    pct(min(dn$share_duplicate))))
}
```

```{r duplication, results="asis"}
# `top` already carries the duplication columns - merging g$duplication in again
# would suffix every one of them .x/.y
if (all(c("dup_runs_tnds", "dup_runs_bods_gtfs") %in% names(top))) {
  dd <- copy(top[, list(zone_id, locality, country, gap, trip_days_tnds,
                        dup_runs_tnds, trip_days_bods_gtfs,
                        dup_runs_bods_gtfs)])
  dd[, share_t := ifelse(trip_days_tnds > 0, dup_runs_tnds / trip_days_tnds, NA)]
  dd[, share_g := ifelse(trip_days_bods_gtfs > 0,
                         dup_runs_bods_gtfs / trip_days_bods_gtfs, NA)]
  cat(sprintf(paste0(
    "Across the %d zones investigated, the duplicate runs still present ",
    "account for a median of **%s** of TNDS's counted trip-days and **%s** of ",
    "the DfT GTFS's. Totals: %s of %s TNDS trip-days and %s of %s BODS GTFS ",
    "trip-days.\n\n"),
    nrow(dd), pct(median(dd$share_t, na.rm = TRUE)),
    pct(median(dd$share_g, na.rm = TRUE)),
    fmt(sum(dd$dup_runs_tnds)), fmt(sum(dd$trip_days_tnds)),
    fmt(sum(dd$dup_runs_bods_gtfs)), fmt(sum(dd$trip_days_bods_gtfs))))

  setorderv(dd, "share_g", -1L)
  # print(): see the note in duplication-national above
  print(knitr::kable(
    data.frame(Zone = dd$zone_id[1:min(12, nrow(dd))],
               Locality = ifelse(is.na(dd$locality), "—", dd$locality)[1:min(12, nrow(dd))],
               `TNDS trip-days` = round(dd$trip_days_tnds)[1:min(12, nrow(dd))],
               `TNDS duplicate` = pct(dd$share_t)[1:min(12, nrow(dd))],
               `BODS trip-days` = round(dd$trip_days_bods_gtfs)[1:min(12, nrow(dd))],
               `BODS duplicate` = pct(dd$share_g)[1:min(12, nrow(dd))],
               check.names = FALSE),
    row.names = FALSE, format.args = list(big.mark = ","),
    caption = "Zones with the largest share of duplicated runs remaining in BODS GTFS"))

  cat(paste0("\nA duplicate here is a statement about the feed, not about the ",
             "road: two identical journeys on one day is one bus described ",
             "twice. These are the ones deduplication would not remove without ",
             "risking real service, so where a source's remaining excess over ",
             "the other is close to its remaining duplicate share, the zone's ",
             "gap is still an artefact of the feed; where it is not, the gap ",
             "is real service one source lacks.\n"))
} else {
  cat("No duplication measurements in this analysis output.\n")
}
```

## What is actually going on in those zones

The route-level breakdown for the largest disagreement in each direction.
"Service" is a group of routes matched across the two sources on route number
and stop pattern, so a service split across several `route_id`s in one source
is compared as one thing.

```{r detail, results="asis"}
# One zone per locality. City-centre LSOAs come in adjacent clusters that all
# contain the same bus station, so the raw top ten is seven versions of one
# place; taking the largest zone of each locality covers distinct places
# instead. Zones with no locality keep their own identity.
distinct_places <- function(d, n) {
  d <- copy(d)
  d[, grp := ifelse(is.na(locality), zone_id, locality)]
  d <- d[, .SD[1], by = grp]
  head(d, n)[, grp := NULL][]
}
detail <- rbind(distinct_places(top[order(-gap)], 5),
                distinct_places(top[order(gap)], 5))
for (i in seq_len(nrow(detail))) {
  d <- detail[i]
  cat(sprintf("\n### %s — %s (%s)\n\n", d$zone_id,
              ifelse(is.na(d$locality), "unnamed", d$locality),
              cn[d$country]))
  cat(sprintf(paste0("TNDS %s trip-runs, BODS GTFS %s, difference **%s**. ",
                     "%d stops in the zone; %d services only in TNDS, %d only ",
                     "in BODS GTFS, %d in both.\n\n"),
              fmt(d$runs_tnds), fmt(d$runs_bods), fmt(d$gap),
              ifelse(is.na(d$stops), 0L, d$stops),
              d$n_only_tnds, d$n_only_bods, d$n_both))
  print(knitr::kable(service_table(d$zone_id), row.names = FALSE,
                     format.args = list(big.mark = ",")))
  cat("\n")
}
```

## Interpretation

```{r interpretation, results="asis"}
big <- top[order(-abs(gap))]
one_sided <- top[runs_bods == 0 | runs_tnds == 0]
cause_tot <- svc[, list(runs = sum(abs(gap)), n = .N), by = cause]
cause_tot <- cause_tot[order(-runs)]

cat(sprintf(paste0(
  "- The largest single-zone disagreement is %s (%s), where the two sources ",
  "differ by %s trip-runs over the four weeks — %s of the larger of the two ",
  "figures.\n"),
  big$zone_id[1], ifelse(is.na(big$locality[1]), "unnamed", big$locality[1]),
  fmt(abs(big$gap[1])), pct(big$rel[1])))

cat(sprintf(paste0(
  "- Within the zones investigated, %s of the difference (by trip-runs) is ",
  "services one source carries and the other does not at all, against %s from ",
  "differing frequencies on shared services.\n"),
  pct(sum(cause_tot$runs[cause_tot$cause != "both, different frequency"]) /
        sum(cause_tot$runs)),
  pct(sum(cause_tot$runs[cause_tot$cause == "both, different frequency"]) /
        sum(cause_tot$runs))))

if (nrow(one_sided) > 0) {
  cat(sprintf(paste0(
    "- %d of the %d zones investigated have **no counted bus service at all** ",
    "in one of the two sources. For those zones the choice of source is not a ",
    "matter of degree: one says the zone has a bus service and the other says ",
    "it has none.\n"), nrow(one_sided), nrow(top)))
}

by_country <- z[, list(zones = .N,
                       disagree = sum(gap != 0),
                       med_abs = median(abs(gap))), by = country]
by_country <- by_country[country %in% names(cn)]
cat("- Disagreement by country:",
    paste(sprintf("%s %s of zones", cn[by_country$country],
                  pct(by_country$disagree / by_country$zones)),
          collapse = "; "), "\n")
```

### Why these particular zones

The zone-level pattern is the national pattern concentrated:

- **Country coverage.** BODS is a statutory requirement for English local bus
  services only. Scottish and Welsh services reach it only where an operator
  crosses the border or publishes voluntarily, so Welsh and Scottish zones
  supply a disproportionate share of the "only in TNDS" cases, up to and
  including zones with no BODS GTFS service at all.
- **Interchanges amplify.** A zone containing a bus station or a major
  interchange has many services calling at many stops, so a single service
  missing from one source moves that zone's total by thousands of runs. The
  largest absolute disagreements are therefore concentrated on town-centre
  zones, and are not evidence that those places are badly described — the
  *relative* column is the fairer reading for them.
- **Snapshot age.** A TNDS snapshot carries the registration operative on the
  day it was taken; the BODS change archive holds future-dated files. Where a
  registration expires inside the counting window, TNDS counts a part-window
  service and BODS GTFS a whole-window one, which appears here as a frequency
  difference on a shared service rather than a missing service. See the
  snapshot-expiry section of the comparison report for the size of this effect.
- **Route naming.** Where the two sources disagree about a service's public
  number — TNDS carrying an operator's marketing name against the DfT's short
  code — the matching links them on stop pattern alone, and only for
  single-source groups. Any pair it still fails to link is counted as exclusive
  to each source, which inflates both "only in" columns at once.
- **Duplicate publication that survives removal.** Measured above, and it works
  in both directions: a source that describes one bus twice reports twice the
  service. Most of it is now removed before counting, but `gtfs_deduplicate()`
  keeps any copy whose dates are not fully covered by the copy kept, or whose
  route number or operator differs — so what is left still moves a zone's
  total. This is one of the few causes on this list that an outside check can
  settle, and where a published timetable has been brought to bear it has
  settled it (see `pdf_validation.md`).

### Caveats

- The unit is trip-runs touching a zone, not passenger-facing frequency. A
  zone crossed by a busy corridor scores highly whether or not the service is
  useful to people living there.
- Zones are the widened LSOA/Data Zone polygons used throughout this analysis
  (small LSOAs unioned with a 500 m buffer of their population-weighted
  centroid, then buffered 100 m). They overlap, so a stop can count in more
  than one zone and the zone totals do not sum to a national trip count.
- The service groupings come from the same route matching the comparison report
  uses, with the same limitations: a wrong pairing shows up here as a spurious
  frequency difference, and a failed pairing as a service missing from both
  directions at once.
- Zones carry only a code, so the "Locality" column is the commonest locality
  prefix among the NaPTAN names of the stops inside the zone. It names the
  place, not the LSOA.
```
