```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.path = "figures/validation-")
library(data.table)
v <- readRDS(validation_path)
lbl <- comparison_source_labels()
```

# Validating the sources against published timetables

The multi-year comparison in `bus_source_comparison.md` can say that TNDS,
BODS TransXChange and BODS GTFS disagree; only an operator's own published
timetable says which of them is right. This checks individual routes against
the documents in `data/example_timetables/`.

Published timetables are easy to obtain for the current week and hard to
obtain for the past, so this uses a **current snapshot** rather than one of
the historic comparison snapshots.

Two sources, not three. BODS TransXChange is excluded here because it is not an
independent answer to the question a document settles: it is the same
operator-published data the DfT's GTFS is made from, differing mainly in which
revision of each dataset survives de-duplication and in who converted it. What a
PDF can adjudicate is TNDS against the DfT's rendering; adding a third column
derived from the same upstream files as the second would not break a tie.
Converting the 1.7 GB change archive also costs hours. The multi-year comparison
covers all three; this does not.

```{r window}
wins <- paste(names(v$window), vapply(v$window, function(x)
  paste(x$startdate, "to", x$enddate), ""), sep = ": ", collapse = "; ")
w <- data.frame(
  Item = c("Snapshot extracted", "Counting windows", "TNDS", "BODS GTFS"),
  Value = c(as.character(v$snapshot$snapshot_date), wins,
            v$snapshot$tnds, v$snapshot$bods_gtfs))
knitr::kable(w)
```

The window opens on the first Monday on or after the extraction date. The
DfT's GTFS carries no history — every calendar in it starts on the extraction
date — so a window opening earlier would silently understate that source.

## Journeys per operating day, published against each source

`Published` is the number of journeys the document prints for one operating
day at the reference stop. Where a document abbreviates a frequent-service
period the block is expanded before counting, and the row is flagged:

* **Expanded** — the count includes journeys recovered from an abbreviation
  such as "at these minutes past each hour until" or "and every 10 mins
  until". These are exact: the printed pattern determines the departures.
* **Upper bound** — the abbreviation was qualified, as in Lothian's "**up
  to** every 10 mins until". The true service is at most this frequent, so
  the expansion is a ceiling and a source counting fewer journeys is not
  necessarily wrong.
* **Reliable** — `FALSE` where any of three checks failed. **Rows marked
  unreliable must not be read as ground truth.** The checks all guard the
  same failure: not a reader that stops, but one that quietly returns a
  plausible subset of the day.
    * the document advertises a frequent-service period and nothing was
      expanded from it — the explicit times either side still parse, so the
      figure looks reasonable while most of the service is missing;
    * a lone two-digit cell appeared with nothing on the page to say whether
      it means a headway or a minute past the hour, readings that differ by
      a factor of several;
    * **Continuous** below is `FALSE`.
* **Continuous** — whether the departures read for a day have a plausible
  shape. A single gap that dwarfs the rest of the day's spacing means a
  block was missed, usually because its heading or stop label is written
  differently from the others. The Fife 34 reads as running 06:10–10:10 and
  then nothing until 18:31.

Where several routes share one printed table — Glasgow's "Service No.: 1 1C
1A 1C …", Fife's bare "91A 91A 90A …" — each column is attributed to its
route from the header row and only the routes being counted are kept.
Without that, another route's journeys are credited to this one, which is
what made the Manchester 142 unusable.

```{r published, results = 'asis'}
pub <- rbindlist(Filter(Negate(is.null), v$published), fill = TRUE)
if (nrow(pub)) {
  p <- pub[, list(`Day type` = daytype, Direction = direction,
                  Published = journeys, Expanded = expanded,
                  `Upper bound` = upper_bound, Continuous = continuous,
                  Reliable = reliable),
           by = list(Route = key)]
  # kable() has to be cat()ed: it is not the last expression in this chunk, so
  # knitr discards its value and the table goes missing from the report - which
  # it silently did, while the prose kept referring to it.
  cat("\n", knitr::kable(p), "\n", sep = "\n")
  bad <- unique(pub[reliable == FALSE]$key)
  if (length(bad)) {
    cat("\n**Unreliable:", paste(bad, collapse = ", "),
        "** - a frequent-service abbreviation in these documents was not",
        "expanded, so the counts are far too low. See the notes below.\n")
  }
} else {
  cat("No timetable could be read.\n")
}
```

## Window totals

Journeys counted over the whole window, per source. These are directly
comparable with the `Published` figures above only after multiplying by the
number of each day type in the window; `Ratio` is TNDS ÷ BODS GTFS, which
needs no such conversion and is where the two sources can be seen to disagree.

A route is located by its public number, its operator, and the stop the
document counts. The stop is what makes the comparison like-for-like. A bus
number is unique only within a town and the operator patterns have to stay
loose, because the sources name the same operator differently - TNDS files the
St Helens 320 under "Arriva Merseyside" where the DfT's GTFS says "Arriva North
West". Number plus operator alone therefore collected eighteen `route_id`s for
the Fife 2A (Oxford, Gloucester, Sheffield, Skegness among them) and nine for
the Falkirk 38 across seven First subsidiaries. Both sources were inflated
equally, so they agreed with each other and diverged wildly from the document -
which reads as a shared converter problem rather than the measurement error it
was. `Stop matched` below is `FALSE` where the reference stop could not be
found in that feed and the count fell back to every same-numbered route of a
matching operator; those rows are the loose comparison and should be read with
that in mind.

```{r feeds}
f <- dcast(v$feeds, key + window ~ source, value.var = "runs", fill = 0)
if (all(c("tnds", "bods_gtfs") %in% names(f))) {
  f[, Ratio := ifelse(bods_gtfs > 0, sprintf("%.2f", tnds / bods_gtfs), "—")]
}
setnames(f, c("key", "window"), c("Route", "Window"))
knitr::kable(f)
```

```{r window-decay, results = 'asis'}
# The second window against the first, per source. Written as computed values
# rather than quoted ones: the trim overrun below is arithmetic, but how far it
# bites depends on the snapshot, so it must not be hard-coded into the prose.
wn <- names(v$window)
trim_days <- tnds_trim_days()
snap <- as.Date(v$snapshot$snapshot_date)
tnds_end <- snap + trim_days
over <- as.integer(v$window[[wn[2]]]$enddate - tnds_end)

d <- dcast(v$feeds, key + source ~ window, value.var = "runs", fill = 0)
d <- d[get(wn[1]) > 0]
d[, ratio := get(wn[2]) / get(wn[1])]
rr <- d[, list(n = .N, med = median(ratio)), by = source]
tn <- rr[source == "tnds"]; bg <- rr[source == "bods_gtfs"]
band <- d[source == "tnds" & ratio > 0.5, list(lo = min(ratio), hi = max(ratio),
                                               n = .N)]
both_low <- intersect(d[source == "tnds" & ratio <= 0.5]$key,
                      d[source == "bods_gtfs" & ratio <= 0.5]$key)

if (over > 0) {
  cat(sprintf(paste0(
    "The two windows are **not** comparable for TNDS, and the table shows why. ",
    "A TNDS snapshot is trimmed to the snapshot date plus or minus %d days by ",
    "`convert_tnds_snapshot()`, so the %s feed ends on **%s** - %d days before ",
    "the `%s` window closes on %s. Those %d days are empty in TNDS by ",
    "construction, which predicts a second-window count of %d/28 = **%.3f** of ",
    "a full one. Measured: the TNDS `%s` divided by `%s` ratio has a median of ",
    "%.3f and sits between %.3f and %.3f for %d of the %d routes, against a ",
    "median of %.3f for BODS GTFS. The apparent collapse of TNDS in the second ",
    "window is therefore an artefact of the conversion trim, not a property of ",
    "the source, and **the `%s` TNDS column must not be read as a service ",
    "level**.\n\n"),
    trim_days, format(snap), format(tnds_end), over, wn[2],
    format(v$window[[wn[2]]]$enddate), over, 28L - over, (28 - over) / 28,
    wn[2], wn[1], tn$med, band$lo, band$hi, band$n, tn$n, bg$med, wn[2]))
} else {
  # The trim used to be 31 days, which stopped 11 days short of this window and
  # made every TNDS figure in it 17/28 of the first window's. tnds_trim_days()
  # is now 45, so what remains here is the source, not the conversion.
  cat(sprintf(paste0(
    "Both windows are now inside what the conversion keeps. ",
    "`convert_tnds_snapshot()` trims to the snapshot date plus or minus %d ",
    "days, so the %s feed runs to **%s**, %d days past the close of the `%s` ",
    "window on %s. An earlier 31-day trim stopped short of this window and ",
    "made every TNDS count in it a fixed 17/28 of the first window's, which is ",
    "no longer the case: the TNDS `%s` divided by `%s` ratio has a median of ",
    "%.3f across %d routes against %.3f for BODS GTFS.\n\n"),
    trim_days, format(snap), format(tnds_end), -over, wn[2],
    format(v$window[[wn[2]]]$enddate), wn[2], wn[1], tn$med, tn$n, bg$med))

  cat(paste0(
    "What is left is a genuine property of a snapshot rather than of the ",
    "conversion: the second window reaches six weeks past the extraction date, ",
    "so a registration that expires in between is carried by the DfT's ",
    "future-dated files and not by TNDS. Read the second window as *service as ",
    "registered on the snapshot date, projected six weeks forward*, and see the ",
    "snapshot-expiry section of `bus_source_comparison.md` for its size.\n\n"))
}

if (length(both_low)) {
  cat(sprintf(paste0(
    "%s fall below half in *both* sources by the same factor, so those are a ",
    "real change in the timetable rather than anything to do with the trim. ",
    "Both are First Glasgow, and the drop falls where a Scottish summer ",
    "timetable would end.\n\n"),
    paste0("`", both_low, "`", collapse = " and ")))
}
```

The second window exists for two bank-holiday references — the Cardiff 62's
"Sundays & public holidays" table and Kinchbus's "Sunday & Bank Holiday
Monday" table. With the trim widened past the end of that window, both are
testable against TNDS for the first time.

### Coverage the DfT feed has since gained

```{r regained, results = 'asis'}
# Several routes were collected precisely because BODS GTFS carried none of
# them - their notes record the February figure as "N against 0". Detect those
# and ask what this snapshot says, rather than leaving the notes to imply a gap
# that may have closed.
w1 <- names(v$window)[1]
# NOT a column called `key`: data.table() reads that as its key= argument and
# the column vanishes
notes <- data.table(route = vapply(v$routes, `[[`, "", "key"),
                    note = vapply(v$routes, function(r)
                      paste(r$note, collapse = " "), ""))
was_zero <- notes[grepl("against 0\\b", note)]$route
fz <- dcast(v$feeds[window == w1 & key %in% was_zero], key ~ source,
            value.var = "runs")
if (nrow(fz)) {
  fz[, verdict := ifelse(bods_gtfs > 0, "now carried", "still absent")]
  print(knitr::kable(
    data.frame(Route = fz$key, TNDS = round(fz$tnds),
               `BODS GTFS` = round(fz$bods_gtfs),
               `Ratio` = ifelse(fz$bods_gtfs > 0,
                                sprintf("%.2f", fz$tnds / fz$bods_gtfs), "—"),
               Verdict = fz$verdict, check.names = FALSE),
    row.names = FALSE, format.args = list(big.mark = ","),
    caption = "Routes selected because BODS GTFS carried nothing for them"))
  n_back <- sum(fz$bods_gtfs > 0)
  if (n_back > 0) {
    r <- fz$tnds[fz$bods_gtfs > 0] / fz$bods_gtfs[fz$bods_gtfs > 0]
    level <- if (all(abs(r - 1) < 0.005)) "at exactly the TNDS level" else
      sprintf("at %s of the TNDS level",
              paste(sprintf("%.2f", r), collapse = ", "))
    cat(sprintf(paste0(
      "\n**%d of these %d routes are now in the DfT's GTFS, %s.** Each was ",
      "collected because that feed carried nothing at all for it in February ",
      "2026, so this is a coverage gap that has closed rather than persisted. ",
      "Two things follow: the February figures inside the notes must not be ",
      "read as current, and a coverage difference measured on one snapshot ",
      "cannot be assumed to hold on another — which is a caution that applies ",
      "to the year-by-year comparison as much as to this report.\n\n"),
      n_back, nrow(fz), level))
  }
}
```

## Which source does the document agree with?

The two tables above compare the sources with each other and with a single
operating day. This one compares each source with the **document's own implied
window total**: each printed day type multiplied by how many days of that type
the window contains, summed. That is the measurement the rest of this report
exists to make, and it says which source is right rather than only that they
differ.

Only the first window, and only routes whose printed tables cover a whole week
(`MF`+`Sa`+`Su`, or `MT`+`Fr`+`Sa`+`Su`) — a route with no Sunday table cannot
be compared against a feed total that includes Sundays. `SuBh` is counted as
Sundays alone: the bank holiday it also covers falls in the second window.

**Read the Edition column first.** A ratio away from 1 says something about a
source only when the document describes the same service the snapshot holds.
Operators leave superseded timetables on their websites, and several of these
documents predate the snapshot by a year or more: where a stale document and
both sources disagree by the same factor, the network changed, and neither
source is at fault. The date is the printed validity statement from the first
page ("From 31 August 2025", "Valid from Monday 24th November 2025"), or failing
that an unambiguous `yyyymmdd` in the file name — a convention this set can
check rather than assume, since the two Cardiff documents named `-20260719-`
also print "19/07/2026" on the page. Blank means neither was found, or the
document is a Word extract rather than a PDF.

```{r published-vs-sources, results = 'asis'}
wmain <- names(v$window)[1]
w1 <- v$window[[wmain]]
# Integer weekdays, not weekdays(): the names that function returns depend on
# the locale, and this must not read differently on another machine.
days <- tabulate(lubridate::wday(seq(w1$startdate, w1$enddate, by = 1),
                                 week_start = 1), 7L)
daytype_days <- list(MT = 1:4, Fr = 5L, MF = 1:5, Sa = 6L, Su = 7L, SuBh = 7L)

p <- rbindlist(Filter(Negate(is.null), v$published), fill = TRUE)
unknown <- setdiff(unique(p$daytype), names(daytype_days))
if (length(unknown)) {
  stop("Unmapped published day type(s): ", paste(unknown, collapse = ", "))
}
# Directions are separate printed tables of one day's service, so they add
p <- p[, list(journeys = sum(journeys), reliable = all(reliable),
              continuous = all(continuous | is.na(continuous))),
       by = list(key, daytype)]
p[, dcount := vapply(daytype, function(x) sum(days[daytype_days[[x]]]), 0)]

per_route <- p[, list(expected = sum(journeys * dcount),
                      days_covered = sum(dcount),
                      reliable = all(reliable) && all(continuous)),
               by = key]
whole_week <- per_route[days_covered == sum(days)]

fm <- dcast(v$feeds[window == wmain], key ~ source, value.var = "runs")
sm <- dcast(v$feeds[window == wmain], key ~ source, value.var = "stop_matched")
setnames(sm, setdiff(names(sm), "key"), paste0("sm_", setdiff(names(sm), "key")))
cmp <- merge(merge(whole_week, fm, by = "key"), sm, by = "key")
cmp[, r_tnds := tnds / expected]
cmp[, r_bods := bods_gtfs / expected]

# The printed validity date, from the first page of the document itself rather
# than from the spec's free-text note, so it cannot drift out of step with the
# file. Knitting happens with the working directory set to reports/.
tt_dir <- if (dir.exists("../data/example_timetables")) {
  "../data/example_timetables"
} else "data/example_timetables"
mon <- paste0("(?:January|February|March|April|May|June|July|August|",
              "September|October|November|December)")
edition_of <- function(files) {
  d <- vapply(files, function(f) {
    p <- file.path(tt_dir, f)
    if (!file.exists(p) || !grepl("\\.pdf$", p, ignore.case = TRUE))
      return(NA_real_)
    tx <- try(pdftools::pdf_text(p)[1], silent = TRUE)
    if (inherits(tx, "try-error")) return(NA_real_)
    ln <- trimws(strsplit(tx, "\n")[[1]])
    hd <- paste(utils::head(ln[nzchar(ln)], 4), collapse = " | ")
    m <- regmatches(hd, regexpr(
      sprintf("\\d{1,2}(?:st|nd|rd|th)? %s (?:19|20)\\d{2}", mon), hd,
      perl = TRUE))
    if (!length(m)) {
      m <- regmatches(hd, regexpr("\\d{1,2}/\\d{1,2}/(?:19|20)\\d{2}", hd,
                                  perl = TRUE))
    }
    if (!length(m)) {
      # Several generators encode the valid-from date in the file name and print
      # no date on the page. The convention is verifiable rather than assumed:
      # the two Cardiff documents named ...-20260719-... also print "19/07/2026".
      # Only the unambiguous 8-digit form is read; "r100-260222.pdf" could be
      # either yymmdd or ddmmyy and is left blank.
      fn <- regmatches(basename(f), regexpr("20\\d{6}", basename(f)))
      if (length(fn)) {
        d8 <- as.Date(fn[1], format = "%Y%m%d")
        if (!is.na(d8)) return(as.numeric(d8))
      }
    }
    if (!length(m)) return(NA_real_)
    # strip the ordinal suffix only where it follows a digit: a blanket
    # gsub("st","") turns "31st August" into "31 Augu"
    as.numeric(lubridate::dmy(gsub("(\\d)(st|nd|rd|th)", "\\1", m[1])))
  }, 0)
  if (all(is.na(d))) NA_real_ else min(d, na.rm = TRUE)  # oldest part governs
}
ed <- vapply(v$routes, function(r) edition_of(r$file), 0)
names(ed) <- vapply(v$routes, `[[`, "", "key")
cmp[, edition := as.Date(ed[key], origin = "1970-01-01")]
cmp[, months_old := round(as.numeric(w1$startdate - edition) / 30.44)]
setorderv(cmp, "r_tnds")

knitr::kable(
  data.frame(Route = cmp$key,
             Edition = ifelse(is.na(cmp$edition), "",
                              format(cmp$edition, "%d %b %Y")),
             `Months old` = ifelse(is.na(cmp$months_old), "",
                                   as.character(cmp$months_old)),
             `Document implies` = round(cmp$expected),
             TNDS = round(cmp$tnds),
             `TNDS ÷ doc` = sprintf("%.2f", cmp$r_tnds),
             `BODS GTFS` = round(cmp$bods_gtfs),
             `BODS ÷ doc` = sprintf("%.2f", cmp$r_bods),
             Reliable = cmp$reliable,
             `Stop matched` = cmp$sm_tnds & cmp$sm_bods_gtfs,
             check.names = FALSE),
  row.names = FALSE, format.args = list(big.mark = ","))
```

```{r published-vs-sources-text, results = 'asis'}
ok <- cmp[reliable == TRUE]
near <- function(r) sum(abs(r - 1) <= 0.10)
cat(sprintf(paste0(
  "Across the %d routes with a whole week of readable tables, TNDS is within ",
  "10%% of the document for **%d** of them and the DfT's GTFS for **%d**. ",
  "Median ratio to the document: TNDS %.2f, BODS GTFS %.2f.\n\n"),
  nrow(ok), near(ok$r_tnds), near(ok$r_bods),
  median(ok$r_tnds), median(ok$r_bods)))

far_t <- ok[abs(r_tnds - 1) > 0.25][order(-abs(r_tnds - 1))]
far_g <- ok[abs(r_bods - 1) > 0.25][order(-abs(r_bods - 1))]
desc <- function(d, col) paste(sprintf("`%s` (%.2f)", d$key, d[[col]]),
                               collapse = ", ")
if (nrow(far_t)) {
  cat(sprintf("Routes where **TNDS** is more than 25%% from the document: %s.\n\n",
              desc(far_t, "r_tnds")))
}
if (nrow(far_g)) {
  cat(sprintf(paste0("Routes where **BODS GTFS** is more than 25%% from the ",
                     "document: %s.\n\n"), desc(far_g, "r_bods")))
}

# Both sources wrong by the SAME factor is a different diagnosis from either
# source being wrong: two independently produced feeds do not err identically,
# so what they share - the document - is the suspect. Two ways that happens, and
# the edition date separates them: the document is out of date, or it is current
# and being misread.
agree_far <- ok[abs(r_tnds - r_bods) <= 0.1 & abs(r_tnds - 1) > 0.25]
stale <- agree_far[!is.na(months_old) & months_old > 6]
misread <- agree_far[is.na(months_old) | months_old <= 6]
if (nrow(stale)) {
  cat(sprintf(paste0(
    "**Stale documents, not source problems:** %s. Both sources agree with ",
    "*each other* to within 10%% and diverge from the document by the same ",
    "factor, and each of these documents predates the snapshot by more than six ",
    "months (%s). The service changed after the timetable was printed; neither ",
    "feed is at fault.\n\n"),
    desc(stale[order(-abs(r_tnds - 1))], "r_tnds"),
    paste(sprintf("`%s` %d months", stale$key, stale$months_old),
          collapse = ", ")))
}
if (nrow(misread)) {
  cat(sprintf(paste0(
    "**Read these as document-reading problems:** %s. The two sources agree ",
    "with each other and diverge from the document by the same factor, but the ",
    "document is current (or states no edition), so what to doubt is the ",
    "reading of it. A ratio near 2.00 in this group is the signature of a table ",
    "that prints one direction where the feeds count both.\n\n"),
    desc(misread[order(-abs(r_tnds - 1))], "r_tnds")))
}

# Where the two sources disagree AND one of them matches the document, the
# document has decided the case - which is the whole purpose of collecting it.
decided <- ok[abs(r_tnds - r_bods) > 0.2 &
                (abs(r_tnds - 1) <= 0.1 | abs(r_bods - 1) <= 0.1)]
if (nrow(decided)) {
  decided[, verdict := ifelse(abs(r_tnds - 1) <= 0.1, "TNDS", "BODS GTFS")]
  cat("Cases the document settles:\n\n")
  for (i in seq_len(nrow(decided))) {
    d <- decided[i]
    age <- if (is.na(d$months_old)) "edition not stated" else
      sprintf("edition %s, %d months before the window",
              format(d$edition, "%d %b %Y"), d$months_old)
    cat(sprintf(paste0("* `%s` — the document implies %s journeys; TNDS %s ",
                       "(%.2f), BODS GTFS %s (%.2f). **%s is right** (%s).\n"),
                d$key, format(round(d$expected), big.mark = ","),
                format(round(d$tnds), big.mark = ","), d$r_tnds,
                format(round(d$bods_gtfs), big.mark = ","), d$r_bods,
                d$verdict, age))
  }
  cat("\n")
}
```

A ratio near 1.00 is strong evidence and a ratio far from it is a lead, not a
verdict: the document may print a different edition from the one the snapshot
carries, and `Reliable = FALSE` rows are excluded from the prose above precisely
because their published figure cannot be trusted. `Stop matched = FALSE` means
the reference stop was not found in one of the feeds, so the count is of every
same-numbered route of a matching operator and will be too high.

### The Bristol 21: the case that set the deduplication defaults

This subsection describes the state of the DfT feed **as published**. Every
figure in the tables above is counted after `UK2GTFS::gtfs_deduplicate()` has
run, so what follows is the problem that stage exists to solve, not a defect in
the numbers reported here.

*Measured directly on the DfT feed named at the top of this report, over the
first window, on 2026-07-30. The figures are quoted rather than computed here
because this report is built from the route-level validation data and not from
the feed itself; `lsoa_disagreement.md` computes the national duplication
figure as part of its own pipeline, so that is the copy that stays current.*

First Bristol's 21 is `route_id` 3700 in the DfT's GTFS. As published, over the
first window, that route_id carried **5,816 trip-days**. Reducing every trip to
the set of (stop, departure time) pairs it makes, and counting distinct
(journey, date) pairs rather than trips, leaves **3,296** — the other 2,520 are
the same bus described twice on the same day. The document implies 3,296 and
TNDS counts 3,296. The same pattern, with the same numbers, appears on
`route_id` 5320 (service 31), which points to one duplicated dataset rather than
a per-route accident.

The 21 is also the route that settled two of `gtfs_deduplicate()`'s defaults.
With `match_block = TRUE` it stayed at 5,816, because the DfT fills `block_id`
with a hash generated per dataset revision and no two copies of a journey ever
agree on it; ignoring `block_id` lands the 21 on 3,296 exactly, and First
Bristol's A1 on 6,916 against a published 6,916. `match_operator = "name"`
settled the 279, whose copies sit under two agency records for one company.
Together they took the national removal rate on this feed from 3.2% to 5.2%.

Duplication is not confined to those. Measured across the whole 2026-07-26 DfT
feed over this window, before deduplication, the same-journey-same-day rate is
several per cent of counted bus runs, concentrated rather than spread: several
`route_id`s are duplicated in their entirety. `lsoa_disagreement.md` carries the
current figure.

A caution about how that is measured, because the obvious version of the test is
wrong. Comparing trips by their (stop, time) signature alone, without regard to
date, returns 41.5% of runs — more than ten times the real figure. GTFS models a
school-term journey and its holiday twin as two trips with identical times and
*complementary* calendars, which is correct modelling and not duplication. Only
trips that run the same journey on the same **date** are counted.

```{r stopmatch}
sm <- dcast(v$feeds[window == names(v$window)[1]], key ~ source,
            value.var = "stop_matched")
setnames(sm, "key", "Route")
knitr::kable(sm, caption = "Stop matched: was the reference stop found?")
```

```{r ids}
ids <- dcast(v$feeds[window == names(v$window)[1]], key ~ source,
             value.var = "route_ids", fill = "")
setnames(ids, "key", "Route")
knitr::kable(ids, caption = "route_id(s) matched in each source")
```

Where a source lists several `route_id`s for one route they are summed: a
source that splits one service across several ids must still be compared as
one service. Several ids is in itself worth noting, since it is how duplicate
publication shows up.

Several ids is not *by itself* evidence of duplication, though, and the four
routes where TNDS reads well above BODS GTFS divide cleanly on that point. For
each of them every trip was reduced to the sorted set of its (stop, departure
time) pairs, and those signatures compared across the route's TNDS `route_id`s
(measured on `tnds_20260726_merged.zip`):

| Route | TNDS ÷ GTFS | TNDS `route_id`s | Trip patterns shared between ids | Reading |
|:------|------------:|-----------------:|:---------------------------------|:--------|
| TBALL | 2.00 | 2 | **148 of 148** | the same timetable published twice |
| TBX38 | 3.00 | 6 | **363 of 688**, as three identical pairs | three registrations, each published twice |
| SC125 | 2.00 | 2 | **0 of 731** | duplication the exact test cannot see |
| CDF62 | 3.86 | 3 | **0 of 147** | not duplication |

For TBALL and TBX38 the TNDS excess is therefore an artefact of duplicate
registration: the exact ratios of 2.00 and 3.00 follow directly from publishing
one timetable twice and one three times over, so **TNDS is overcounting them**.

The SC125 row needs correcting, and the correction is the reason
`near_duplicate_journeys.md` exists. "0 of 731" is what the exact test found and
it is accurate, but the conclusion drawn from it — that the 125 is not
duplicated — was wrong. Pair the two `route_id`s on their **stop sequence**
instead of on stops-and-times, and 230 pairs of trips run the same stops on the
same day within one minute of each other at every call, 358 within two minutes.
The service is registered twice by two Stagecoach subsidiaries, `SCCU`
(Stagecoach North West) and `SCMY` (Stagecoach Cumbria and Lancashire), and the
two registrations were written from different working timetables, so not one
journey matches exactly. TBX38 has the same thing on top of its exact
duplication: pairing its six `route_id`s the same way finds 254 pairs at zero
difference and a further 238 within a minute, so the exact test catches part of
its duplication and leaves the rest. See `near_duplicate_journeys.md` for how
common this is nationally and why the tolerance that would catch it cannot
safely be made the default.

That is as far as the trip-pattern evidence goes, and it is worth being precise
about what it does *not* establish. Compared with the documents in the section
below, TBALL's doubled TNDS figure lands nearer the document (1.12) than BODS
GTFS does (0.56), and TBX38 exceeds it in both sources. Two things make that row
unsafe to adjudicate rather than a contradiction: the reference stop is not found
in either feed for these routes, so both counts are of every same-numbered route
of a matching operator, and both documents are Word extracts whose readers are
the least trustworthy in the set (TBALL's Monday–Friday column includes
Fridays-only journeys, so its implied total is itself an overestimate). The
duplication in TNDS is established; that BODS GTFS is therefore *correct* for
these two is not.

CDF62 is the one of the four whose `route_id`s really do carry distinct trips,
and the documents settle it outright — TNDS reads 1.02 of the Cardiff 62's
published timetable and BODS GTFS 0.26, so the missing service is in BODS GTFS,
which is what England-only statutory coverage predicts for a Cardiff Bus
service. SC125 is settled the other way and by the same evidence: TNDS reads
2.28 of the document against BODS GTFS's 1.14, exactly what a service counted
twice looks like, and the near-duplicate pairing above supplies the mechanism.

## Notes on each document

Figures quoted *inside* these notes are **selection rationale**, not
measurements: they come from the February 2026 comparison window that these
documents were picked from. That window has since been dropped from the
comparison (its TNDS side is distorted by registrations expiring inside it), and
in any case it is not the window counted here. Where a note disagrees with the
Window totals table above, **the table is the measurement** — the Wigan `AN320`
is noted as TNDS reading 1.96× BODS GTFS, but on this July snapshot the two
sources agree exactly.

```{r notes}
n <- rbindlist(lapply(v$routes, function(r)
  data.table(Route = r$key,
             Operator = r$operator,
             Skipped = isTRUE(r$skip),
             Note = r$note)))
knitr::kable(n)
```
