```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.path = "figures/validation-")
library(data.table)
v <- readRDS(validation_path)
lbl <- comparison_source_labels()
```

# Validating the sources against published timetables

The multi-year comparison in `bus_source_comparison.md` can say that TNDS,
BODS TransXChange and BODS GTFS disagree; only an operator's own published
timetable says which of them is right. This checks individual routes against
the documents in `data/example_timetables/`.

Published timetables are easy to obtain for the current week and hard to
obtain for the past, so this uses a **current snapshot** rather than one of
the historic comparison snapshots.

Two sources, not three. BODS TransXChange is excluded here: the multi-year
comparison already establishes that a fraction of the operator-published
files fail to convert, so it carries a coverage gap the other two do not, and
checking it against the PDFs would re-measure that gap rather than say
anything new about TNDS or the DfT's GTFS.

```{r window}
wins <- paste(names(v$window), vapply(v$window, function(x)
  paste(x$startdate, "to", x$enddate), ""), sep = ": ", collapse = "; ")
w <- data.frame(
  Item = c("Snapshot extracted", "Counting windows", "TNDS", "BODS GTFS"),
  Value = c(as.character(v$snapshot$snapshot_date), wins,
            v$snapshot$tnds, v$snapshot$bods_gtfs))
knitr::kable(w)
```

The window opens on the first Monday on or after the extraction date. The
DfT's GTFS carries no history — every calendar in it starts on the extraction
date — so a window opening earlier would silently understate that source, as
it did in the February 2026 comparison.

## Journeys per operating day, published against each source

`Published` is the number of journeys the document prints for one operating
day at the reference stop. Where a document abbreviates a frequent-service
period the block is expanded before counting, and the row is flagged:

* **Expanded** — the count includes journeys recovered from an abbreviation
  such as "at these minutes past each hour until" or "and every 10 mins
  until". These are exact: the printed pattern determines the departures.
* **Upper bound** — the abbreviation was qualified, as in Lothian's "**up
  to** every 10 mins until". The true service is at most this frequent, so
  the expansion is a ceiling and a source counting fewer journeys is not
  necessarily wrong.
* **Reliable** — `FALSE` where any of three checks failed. **Rows marked
  unreliable must not be read as ground truth.** The checks all guard the
  same failure: not a reader that stops, but one that quietly returns a
  plausible subset of the day.
    * the document advertises a frequent-service period and nothing was
      expanded from it — the explicit times either side still parse, so the
      figure looks reasonable while most of the service is missing;
    * a lone two-digit cell appeared with nothing on the page to say whether
      it means a headway or a minute past the hour, readings that differ by
      a factor of several;
    * **Continuous** below is `FALSE`.
* **Continuous** — whether the departures read for a day have a plausible
  shape. A single gap that dwarfs the rest of the day's spacing means a
  block was missed, usually because its heading or stop label is written
  differently from the others. The Fife 34 reads as running 06:10–10:10 and
  then nothing until 18:31.

Where several routes share one printed table — Glasgow's "Service No.: 1 1C
1A 1C …", Fife's bare "91A 91A 90A …" — each column is attributed to its
route from the header row and only the routes being counted are kept.
Without that, another route's journeys are credited to this one, which is
what made the Manchester 142 unusable.

```{r published, results = 'asis'}
pub <- rbindlist(Filter(Negate(is.null), v$published), fill = TRUE)
if (nrow(pub)) {
  p <- pub[, list(`Day type` = daytype, Direction = direction,
                  Published = journeys, Expanded = expanded,
                  `Upper bound` = upper_bound, Continuous = continuous,
                  Reliable = reliable),
           by = list(Route = key)]
  # kable() has to be cat()ed: it is not the last expression in this chunk, so
  # knitr discards its value and the table goes missing from the report - which
  # it silently did, while the prose kept referring to it.
  cat("\n", knitr::kable(p), "\n", sep = "\n")
  bad <- unique(pub[reliable == FALSE]$key)
  if (length(bad)) {
    cat("\n**Unreliable:", paste(bad, collapse = ", "),
        "** - a frequent-service abbreviation in these documents was not",
        "expanded, so the counts are far too low. See the notes below.\n")
  }
} else {
  cat("No timetable could be read.\n")
}
```

## Window totals

Journeys counted over the whole window, per source. These are directly
comparable with the `Published` figures above only after multiplying by the
number of each day type in the window, so the ratio column is the useful one.

A route is located by its public number, its operator, and the stop the
document counts. The stop is what makes the comparison like-for-like. A bus
number is unique only within a town and the operator patterns have to stay
loose, because the sources name the same operator differently - TNDS files the
St Helens 320 under "Arriva Merseyside" where the DfT's GTFS says "Arriva North
West". Number plus operator alone therefore collected eighteen `route_id`s for
the Fife 2A (Oxford, Gloucester, Sheffield, Skegness among them) and nine for
the Falkirk 38 across seven First subsidiaries. Both sources were inflated
equally, so they agreed with each other and diverged wildly from the document -
which reads as a shared converter problem rather than the measurement error it
was. `Stop matched` below is `FALSE` where the reference stop could not be
found in that feed and the count fell back to every same-numbered route of a
matching operator; those rows are the loose comparison and should be read with
that in mind.

```{r feeds}
f <- dcast(v$feeds, key + window ~ source, value.var = "runs", fill = 0)
setnames(f, c("key", "window"), c("Route", "Window"))
knitr::kable(f)
```

```{r stopmatch}
sm <- dcast(v$feeds[window == names(v$window)[1]], key ~ source,
            value.var = "stop_matched")
setnames(sm, "key", "Route")
knitr::kable(sm, caption = "Stop matched: was the reference stop found?")
```

```{r ids}
ids <- dcast(v$feeds[window == names(v$window)[1]], key ~ source,
             value.var = "route_ids", fill = "")
setnames(ids, "key", "Route")
knitr::kable(ids, caption = "route_id(s) matched in each source")
```

Where a source lists several `route_id`s for one route they are summed: a
source that splits one service across several ids must still be compared as
one service. Several ids is in itself worth noting, since it is how duplicate
publication shows up.

## Notes on each document

```{r notes}
n <- rbindlist(lapply(v$routes, function(r)
  data.table(Route = r$key,
             Operator = r$operator,
             Skipped = isTRUE(r$skip),
             Note = r$note)))
knitr::kable(n)
```
