```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.path = "figures/validation-")
library(data.table)
v <- readRDS(validation_path)
lbl <- comparison_source_labels()
```

# Validating the sources against published timetables

The multi-year comparison in `bus_source_comparison.md` can say that TNDS,
BODS TransXChange and BODS GTFS disagree; only an operator's own published
timetable says which of them is right. This checks individual routes against
the documents in `data/example_timetables/`.
The section "The zones where TNDS and BODS GTFS disagree most", near the end, turns the
same documents on the 60 zones of `lsoa_disagreement.md` and says, area by
area, which source the operators' timetables support.

Published timetables are easy to obtain for the current week and hard to
obtain for the past, so this uses a **current snapshot** rather than one of
the historic comparison snapshots.

Two sources, not three. BODS TransXChange is excluded here because it is not an
independent answer to the question a document settles: it is the same
operator-published data the DfT's GTFS is made from, differing mainly in which
revision of each dataset survives de-duplication and in who converted it. What a
PDF can adjudicate is TNDS against the DfT's rendering; adding a third column
derived from the same upstream files as the second would not break a tie.
Converting the 1.7 GB change archive also costs hours. The multi-year comparison
covers all three; this does not.

```{r window}
wins <- paste(names(v$window), vapply(v$window, function(x)
  paste(x$startdate, "to", x$enddate), ""), sep = ": ", collapse = "; ")
w <- data.frame(
  Item = c("Snapshot extracted", "Counting windows", "TNDS", "BODS GTFS"),
  Value = c(as.character(v$snapshot$snapshot_date), wins,
            v$snapshot$tnds, v$snapshot$bods_gtfs))
knitr::kable(w)
```

The window opens on the first Monday on or after the extraction date. The
DfT's GTFS carries no history — every calendar in it starts on the extraction
date — so a window opening earlier would silently understate that source.

## Journeys per operating day, published against each source

`Published` is the number of journeys the document prints for one operating
day at the reference stop. Where a document abbreviates a frequent-service
period the block is expanded before counting, and the row is flagged:

* **Expanded** — the count includes journeys recovered from an abbreviation
  such as "at these minutes past each hour until" or "and every 10 mins
  until". These are exact: the printed pattern determines the departures.
* **Upper bound** — the abbreviation was qualified, as in Lothian's "**up
  to** every 10 mins until". The true service is at most this frequent, so
  the expansion is a ceiling and a source counting fewer journeys is not
  necessarily wrong.
* **Reliable** — `FALSE` where any of three checks failed. **Rows marked
  unreliable must not be read as ground truth.** The checks all guard the
  same failure: not a reader that stops, but one that quietly returns a
  plausible subset of the day.
    * the document advertises a frequent-service period and nothing was
      expanded from it — the explicit times either side still parse, so the
      figure looks reasonable while most of the service is missing;
    * a lone two-digit cell appeared with nothing on the page to say whether
      it means a headway or a minute past the hour, readings that differ by
      a factor of several;
    * **Continuous** below is `FALSE`.
* **Continuous** — whether the departures read for a day have a plausible
  shape. A single gap that dwarfs the rest of the day's spacing means a
  block was missed, usually because its heading or stop label is written
  differently from the others. The Fife 34 reads as running 06:10–10:10 and
  then nothing until 18:31.

Where several routes share one printed table — Glasgow's "Service No.: 1 1C
1A 1C …", Fife's bare "91A 91A 90A …" — each column is attributed to its
route from the header row and only the routes being counted are kept.
Without that, another route's journeys are credited to this one, which is
what made the Manchester 142 unusable.

```{r published, results = 'asis'}
pub <- rbindlist(Filter(Negate(is.null), v$published), fill = TRUE)
if (nrow(pub)) {
  p <- pub[, list(`Day type` = daytype, Direction = direction,
                  Published = journeys, Expanded = expanded,
                  `Upper bound` = upper_bound, Continuous = continuous,
                  Reliable = reliable),
           by = list(Route = key)]
  # kable() has to be cat()ed: it is not the last expression in this chunk, so
  # knitr discards its value and the table goes missing from the report - which
  # it silently did, while the prose kept referring to it.
  cat("\n", knitr::kable(p), "\n", sep = "\n")
  bad <- unique(pub[reliable == FALSE]$key)
  if (length(bad)) {
    cat("\n**Unreliable:", paste(bad, collapse = ", "),
        "** - a frequent-service abbreviation in these documents was not",
        "expanded, so the counts are far too low. See the notes below.\n")
  }
} else {
  cat("No timetable could be read.\n")
}
```

## Window totals

Journeys counted over the whole window, per source. These are directly
comparable with the `Published` figures above only after multiplying by the
number of each day type in the window; `Ratio` is TNDS ÷ BODS GTFS, which
needs no such conversion and is where the two sources can be seen to disagree.

A route is located by its public number, its operator, and the stop the
document counts. The stop is what makes the comparison like-for-like. A bus
number is unique only within a town and the operator patterns have to stay
loose, because the sources name the same operator differently - TNDS files the
St Helens 320 under "Arriva Merseyside" where the DfT's GTFS says "Arriva North
West". Number plus operator alone therefore collected eighteen `route_id`s for
the Fife 2A (Oxford, Gloucester, Sheffield, Skegness among them) and nine for
the Falkirk 38 across seven First subsidiaries. Both sources were inflated
equally, so they agreed with each other and diverged wildly from the document -
which reads as a shared converter problem rather than the measurement error it
was. `Stop matched` below is `FALSE` where the reference stop could not be
found in that feed and the count fell back to every same-numbered route of a
matching operator; those rows are the loose comparison and should be read with
that in mind.

```{r feeds}
f <- dcast(v$feeds, key + window ~ source, value.var = "runs", fill = 0)
if (all(c("tnds", "bods_gtfs") %in% names(f))) {
  f[, Ratio := ifelse(bods_gtfs > 0, sprintf("%.2f", tnds / bods_gtfs), "—")]
}
setnames(f, c("key", "window"), c("Route", "Window"))
knitr::kable(f)
```

```{r window-decay, results = 'asis'}
# The second window against the first, per source. Written as computed values
# rather than quoted ones: the trim overrun below is arithmetic, but how far it
# bites depends on the snapshot, so it must not be hard-coded into the prose.
wn <- names(v$window)
trim_days <- tnds_trim_days()
snap <- as.Date(v$snapshot$snapshot_date)
tnds_end <- snap + trim_days
over <- as.integer(v$window[[wn[2]]]$enddate - tnds_end)
# Window length is read off the window, not written in: it was 28 days until
# October 2026 and is 14 now (study_weeks()), and the trim-shortfall
# arithmetic below is in units of it.
wlen <- as.integer(v$window[[wn[2]]]$enddate -
                     v$window[[wn[2]]]$startdate) + 1L

d <- dcast(v$feeds, key + source ~ window, value.var = "runs", fill = 0)
d <- d[get(wn[1]) > 0]
d[, ratio := get(wn[2]) / get(wn[1])]
rr <- d[, list(n = .N, med = median(ratio)), by = source]
tn <- rr[source == "tnds"]; bg <- rr[source == "bods_gtfs"]
band <- d[source == "tnds" & ratio > 0.5, list(lo = min(ratio), hi = max(ratio),
                                               n = .N)]
both_low <- intersect(d[source == "tnds" & ratio <= 0.5]$key,
                      d[source == "bods_gtfs" & ratio <= 0.5]$key)

if (over > 0) {
  cat(sprintf(paste0(
    "The two windows are **not** comparable for TNDS, and the table shows why. ",
    "A TNDS snapshot is trimmed to the snapshot date plus or minus %d days by ",
    "`convert_tnds_snapshot()`, so the %s feed ends on **%s** - %d days before ",
    "the `%s` window closes on %s. Those %d days are empty in TNDS by ",
    "construction, which predicts a second-window count of %d/%d = **%.3f** of ",
    "a full one. Measured: the TNDS `%s` divided by `%s` ratio has a median of ",
    "%.3f and sits between %.3f and %.3f for %d of the %d routes, against a ",
    "median of %.3f for BODS GTFS. The apparent collapse of TNDS in the second ",
    "window is therefore an artefact of the conversion trim, not a property of ",
    "the source, and **the `%s` TNDS column must not be read as a service ",
    "level**.\n\n"),
    trim_days, format(snap), format(tnds_end), over, wn[2],
    format(v$window[[wn[2]]]$enddate), over, wlen - over, wlen,
    (wlen - over) / wlen,
    wn[2], wn[1], tn$med, band$lo, band$hi, band$n, tn$n, bg$med, wn[2]))
} else {
  # The trim used to be 31 days, which stopped 11 days short of this window
  # and made every TNDS figure in it 17/28 of the first window's (the windows
  # were 28 days then). tnds_trim_days() is now 45, so what remains here is
  # the source, not the conversion.
  cat(sprintf(paste0(
    "Both windows are now inside what the conversion keeps. ",
    "`convert_tnds_snapshot()` trims to the snapshot date plus or minus %d ",
    "days, so the %s feed runs to **%s**, %d days past the close of the `%s` ",
    "window on %s. An earlier 31-day trim stopped short of this window and ",
    "made every TNDS count in it a fixed 17/28 of the first window's, which is ",
    "no longer the case: the TNDS `%s` divided by `%s` ratio has a median of ",
    "%.3f across %d routes against %.3f for BODS GTFS.\n\n"),
    trim_days, format(snap), format(tnds_end), -over, wn[2],
    format(v$window[[wn[2]]]$enddate), wn[2], wn[1], tn$med, tn$n, bg$med))

  cat(paste0(
    "What is left is a genuine property of a snapshot rather than of the ",
    "conversion: the second window reaches six weeks past the extraction date, ",
    "so a registration that expires in between is carried by the DfT's ",
    "future-dated files and not by TNDS. Read the second window as *service as ",
    "registered on the snapshot date, projected six weeks forward*, and see the ",
    "snapshot-expiry section of `bus_source_comparison.md` for its size.\n\n"))
}

if (length(both_low)) {
  cat(sprintf(paste0(
    "%s fall below half in *both* sources by the same factor, so those are a ",
    "real change in the timetable rather than anything to do with the trim. ",
    "Both are First Glasgow, and the drop falls where a Scottish summer ",
    "timetable would end.\n\n"),
    paste0("`", both_low, "`", collapse = " and ")))
}
```

The second window exists for two bank-holiday references — the Cardiff 62's
"Sundays & public holidays" table and Kinchbus's "Sunday & Bank Holiday
Monday" table. With the trim widened past the end of that window, both are
testable against TNDS for the first time.

### Coverage the DfT feed has since gained

```{r regained, results = 'asis'}
# Several routes were collected precisely because BODS GTFS carried none of
# them - their notes record the February figure as "N against 0". Detect those
# and ask what this snapshot says, rather than leaving the notes to imply a gap
# that may have closed.
w1 <- names(v$window)[1]
# NOT a column called `key`: data.table() reads that as its key= argument and
# the column vanishes
notes <- data.table(route = vapply(v$routes, `[[`, "", "key"),
                    note = vapply(v$routes, function(r)
                      paste(r$note, collapse = " "), ""))
was_zero <- notes[grepl("against 0\\b", note)]$route
fz <- dcast(v$feeds[window == w1 & key %in% was_zero], key ~ source,
            value.var = "runs")
if (nrow(fz)) {
  fz[, verdict := ifelse(bods_gtfs > 0, "now carried", "still absent")]
  print(knitr::kable(
    data.frame(Route = fz$key, TNDS = round(fz$tnds),
               `BODS GTFS` = round(fz$bods_gtfs),
               `Ratio` = ifelse(fz$bods_gtfs > 0,
                                sprintf("%.2f", fz$tnds / fz$bods_gtfs), "—"),
               Verdict = fz$verdict, check.names = FALSE),
    row.names = FALSE, format.args = list(big.mark = ","),
    caption = "Routes selected because BODS GTFS carried nothing for them"))
  n_back <- sum(fz$bods_gtfs > 0)
  if (n_back > 0) {
    r <- fz$tnds[fz$bods_gtfs > 0] / fz$bods_gtfs[fz$bods_gtfs > 0]
    level <- if (all(abs(r - 1) < 0.005)) "at exactly the TNDS level" else
      sprintf("at %s of the TNDS level",
              paste(sprintf("%.2f", r), collapse = ", "))
    cat(sprintf(paste0(
      "\n**%d of these %d routes are now in the DfT's GTFS, %s.** Each was ",
      "collected because that feed carried nothing at all for it in February ",
      "2026, so this is a coverage gap that has closed rather than persisted. ",
      "Two things follow: the February figures inside the notes must not be ",
      "read as current, and a coverage difference measured on one snapshot ",
      "cannot be assumed to hold on another — which is a caution that applies ",
      "to the year-by-year comparison as much as to this report.\n\n"),
      n_back, nrow(fz), level))
  }
}
```

## Which source does the document agree with?

The two tables above compare the sources with each other and with a single
operating day. This one compares each source with the **document's own implied
window total**: each printed day type multiplied by how many days of that type
the window contains, summed. That is the measurement the rest of this report
exists to make, and it says which source is right rather than only that they
differ.

Only the first window, and only routes whose printed tables cover a whole week
(`MF`+`Sa`+`Su`, or `MT`+`Fr`+`Sa`+`Su`) — a route with no Sunday table cannot
be compared against a feed total that includes Sundays. `SuBh` is counted as
Sundays alone: the bank holiday it also covers falls in the second window.

**Read the Edition column first.** A ratio away from 1 says something about a
source only when the document describes the same service the snapshot holds.
Operators leave superseded timetables on their websites, and several of these
documents predate the snapshot by a year or more: where a stale document and
both sources disagree by the same factor, the network changed, and neither
source is at fault. The date is the printed validity statement from the first
page ("From 31 August 2025", "Valid from Monday 24th November 2025"), or failing
that an unambiguous `yyyymmdd` in the file name — a convention this set can
check rather than assume, since the two Cardiff documents named `-20260719-`
also print "19/07/2026" on the page. Blank means neither was found, or the
document is a Word extract rather than a PDF.

```{r published-vs-sources, results = 'asis'}
wmain <- names(v$window)[1]
w1 <- v$window[[wmain]]
# Integer weekdays, not weekdays(): the names that function returns depend on
# the locale, and this must not read differently on another machine.
days <- tabulate(lubridate::wday(seq(w1$startdate, w1$enddate, by = 1),
                                 week_start = 1), 7L)
daytype_days <- list(MT = 1:4, Fr = 5L, MF = 1:5, Sa = 6L, Su = 7L, SuBh = 7L)

p <- rbindlist(Filter(Negate(is.null), v$published), fill = TRUE)
unknown <- setdiff(unique(p$daytype), names(daytype_days))
if (length(unknown)) {
  stop("Unmapped published day type(s): ", paste(unknown, collapse = ", "))
}
# Directions are separate printed tables of one day's service, so they add
p <- p[, list(journeys = sum(journeys), reliable = all(reliable),
              continuous = all(continuous | is.na(continuous))),
       by = list(key, daytype)]
p[, dcount := vapply(daytype, function(x) sum(days[daytype_days[[x]]]), 0)]

per_route <- p[, list(expected = sum(journeys * dcount),
                      days_covered = sum(dcount),
                      reliable = all(reliable) && all(continuous)),
               by = key]
whole_week <- per_route[days_covered == sum(days)]

fm <- dcast(v$feeds[window == wmain], key ~ source, value.var = "runs")
sm <- dcast(v$feeds[window == wmain], key ~ source, value.var = "stop_matched")
setnames(sm, setdiff(names(sm), "key"), paste0("sm_", setdiff(names(sm), "key")))
cmp <- merge(merge(whole_week, fm, by = "key"), sm, by = "key")
cmp[, r_tnds := tnds / expected]
cmp[, r_bods := bods_gtfs / expected]

# The printed validity date, from the first page of the document itself rather
# than from the spec's free-text note, so it cannot drift out of step with the
# file. Knitting happens with the working directory set to reports/.
tt_dir <- if (dir.exists("../data/example_timetables")) {
  "../data/example_timetables"
} else "data/example_timetables"
mon <- paste0("(?:January|February|March|April|May|June|July|August|",
              "September|October|November|December)")
edition_of <- function(files) {
  d <- vapply(files, function(f) {
    p <- file.path(tt_dir, f)
    if (!file.exists(p) || !grepl("\\.pdf$", p, ignore.case = TRUE))
      return(NA_real_)
    tx <- try(pdftools::pdf_text(p)[1], silent = TRUE)
    if (inherits(tx, "try-error")) return(NA_real_)
    ln <- trimws(strsplit(tx, "\n")[[1]])
    hd <- paste(utils::head(ln[nzchar(ln)], 4), collapse = " | ")
    m <- regmatches(hd, regexpr(
      sprintf("\\d{1,2}(?:st|nd|rd|th)? %s (?:19|20)\\d{2}", mon), hd,
      perl = TRUE))
    if (!length(m)) {
      m <- regmatches(hd, regexpr("\\d{1,2}/\\d{1,2}/(?:19|20)\\d{2}", hd,
                                  perl = TRUE))
    }
    if (!length(m)) {
      # Several generators encode the valid-from date in the file name and print
      # no date on the page. The convention is verifiable rather than assumed:
      # the two Cardiff documents named ...-20260719-... also print "19/07/2026".
      # Only the unambiguous 8-digit form is read; "r100-260222.pdf" could be
      # either yymmdd or ddmmyy and is left blank.
      fn <- regmatches(basename(f), regexpr("20\\d{6}", basename(f)))
      if (length(fn)) {
        d8 <- as.Date(fn[1], format = "%Y%m%d")
        if (!is.na(d8)) return(as.numeric(d8))
      }
    }
    if (!length(m)) return(NA_real_)
    # strip the ordinal suffix only where it follows a digit: a blanket
    # gsub("st","") turns "31st August" into "31 Augu"
    as.numeric(lubridate::dmy(gsub("(\\d)(st|nd|rd|th)", "\\1", m[1])))
  }, 0)
  if (all(is.na(d))) NA_real_ else min(d, na.rm = TRUE)  # oldest part governs
}
ed <- vapply(v$routes, function(r) edition_of(r$file), 0)
names(ed) <- vapply(v$routes, `[[`, "", "key")
cmp[, edition := as.Date(ed[key], origin = "1970-01-01")]
cmp[, months_old := round(as.numeric(w1$startdate - edition) / 30.44)]
setorderv(cmp, "r_tnds")

knitr::kable(
  data.frame(Route = cmp$key,
             Edition = ifelse(is.na(cmp$edition), "",
                              format(cmp$edition, "%d %b %Y")),
             `Months old` = ifelse(is.na(cmp$months_old), "",
                                   as.character(cmp$months_old)),
             `Document implies` = round(cmp$expected),
             TNDS = round(cmp$tnds),
             `TNDS ÷ doc` = sprintf("%.2f", cmp$r_tnds),
             `BODS GTFS` = round(cmp$bods_gtfs),
             `BODS ÷ doc` = sprintf("%.2f", cmp$r_bods),
             Reliable = cmp$reliable,
             `Stop matched` = cmp$sm_tnds & cmp$sm_bods_gtfs,
             check.names = FALSE),
  row.names = FALSE, format.args = list(big.mark = ","))
```

```{r published-vs-sources-text, results = 'asis'}
ok <- cmp[reliable == TRUE]
near <- function(r) sum(abs(r - 1) <= 0.10)
cat(sprintf(paste0(
  "Across the %d routes with a whole week of readable tables, TNDS is within ",
  "10%% of the document for **%d** of them and the DfT's GTFS for **%d**. ",
  "Median ratio to the document: TNDS %.2f, BODS GTFS %.2f.\n\n"),
  nrow(ok), near(ok$r_tnds), near(ok$r_bods),
  median(ok$r_tnds), median(ok$r_bods)))

far_t <- ok[abs(r_tnds - 1) > 0.25][order(-abs(r_tnds - 1))]
far_g <- ok[abs(r_bods - 1) > 0.25][order(-abs(r_bods - 1))]
desc <- function(d, col) paste(sprintf("`%s` (%.2f)", d$key, d[[col]]),
                               collapse = ", ")
if (nrow(far_t)) {
  cat(sprintf("Routes where **TNDS** is more than 25%% from the document: %s.\n\n",
              desc(far_t, "r_tnds")))
}
if (nrow(far_g)) {
  cat(sprintf(paste0("Routes where **BODS GTFS** is more than 25%% from the ",
                     "document: %s.\n\n"), desc(far_g, "r_bods")))
}

# Both sources wrong by the SAME factor is a different diagnosis from either
# source being wrong: two independently produced feeds do not err identically,
# so what they share - the document - is the suspect. Two ways that happens, and
# the edition date separates them: the document is out of date, or it is current
# and being misread.
agree_far <- ok[abs(r_tnds - r_bods) <= 0.1 & abs(r_tnds - 1) > 0.25]
stale <- agree_far[!is.na(months_old) & months_old > 6]
misread <- agree_far[is.na(months_old) | months_old <= 6]
if (nrow(stale)) {
  cat(sprintf(paste0(
    "**Stale documents, not source problems:** %s. Both sources agree with ",
    "*each other* to within 10%% and diverge from the document by the same ",
    "factor, and each of these documents predates the snapshot by more than six ",
    "months (%s). The service changed after the timetable was printed; neither ",
    "feed is at fault.\n\n"),
    desc(stale[order(-abs(r_tnds - 1))], "r_tnds"),
    paste(sprintf("`%s` %d months", stale$key, stale$months_old),
          collapse = ", ")))
}
if (nrow(misread)) {
  cat(sprintf(paste0(
    "**Read these as document-reading problems:** %s. The two sources agree ",
    "with each other and diverge from the document by the same factor, but the ",
    "document is current (or states no edition), so what to doubt is the ",
    "reading of it. A ratio near 2.00 in this group is the signature of a table ",
    "that prints one direction where the feeds count both.\n\n"),
    desc(misread[order(-abs(r_tnds - 1))], "r_tnds")))
}

# The doubling group: BODS GTFS half again the document or more while TNDS
# stays near it. The shape is the same as the zone-level gap in
# lsoa_disagreement.md, so the two analyses are seeing one mechanism from
# opposite ends. Graded rather than pooled - a route only carries full weight
# if the document is current, the week is complete AND the reference stop was
# found in both feeds, and lumping the weaker rows in with those would
# overstate how much of this rests on evidence that can bear it.
shape <- cmp[sm_tnds & sm_bods_gtfs & r_bods >= 1.5 & r_tnds <= 1.5 &
               r_bods / r_tnds >= 1.4]
shape[, core := reliable & !is.na(months_old) & months_old <= 6]
data.table::setorderv(shape, c("core", "r_bods"), c(-1L, -1L))
dbl <- shape[core == TRUE]
sup <- shape[core == FALSE]
row_line <- function(d, why = NULL) {
  cat(sprintf("* `%s` — document %s; TNDS %s (%.2f), BODS GTFS %s (%.2f)%s.\n",
              d$key, format(round(d$expected), big.mark = ","),
              format(round(d$tnds), big.mark = ","), d$r_tnds,
              format(round(d$bods_gtfs), big.mark = ","), d$r_bods,
              if (is.null(why)) "" else paste0(" — ", why)))
}
if (nrow(dbl)) {
  cat(sprintf(paste0(
    "**The doubling group.** %d route%s clear%s every bar at once: an edition ",
    "printed for the week being counted (six months old or less), a complete ",
    "week of readable tables, the reference stop found in both feeds, and the ",
    "DfT's GTFS reading at least half again the document while TNDS stays ",
    "near it.\n\n"),
    nrow(dbl), if (nrow(dbl) == 1) "" else "s",
    if (nrow(dbl) == 1) "s" else ""))
  for (i in seq_len(nrow(dbl))) row_line(dbl[i])
  cat("\n")
}
if (nrow(sup)) {
  cat(sprintf(paste0(
    "%s route%s show%s the same shape on weaker evidence. %s not ",
    "independent confirmation - each fails a bar the group above clears - but ",
    "%s fails a *different* bar, and a reading error in the DfT's feed is a ",
    "simpler explanation of all %d than %d unrelated faults that happen to ",
    "point the same way.\n\n"),
    nrow(sup), if (nrow(sup) == 1) "" else "s",
    if (nrow(sup) == 1) "s" else "",
    if (nrow(sup) == 1) "It is" else "They are",
    if (nrow(sup) == 1) "it" else "each",
    nrow(sup) + nrow(dbl), nrow(sup)))
  for (i in seq_len(nrow(sup))) {
    d <- sup[i]
    row_line(d, if (!d$reliable) {
      "the week of tables is incomplete"
    } else if (is.na(d$months_old)) {
      "the document states no edition"
    } else {
      sprintf("the document is %d months old", d$months_old)
    })
  }
  cat("\n")
}

# Where the two sources disagree AND one of them matches the document, the
# document has decided the case - which is the whole purpose of collecting it.
decided <- ok[abs(r_tnds - r_bods) > 0.2 &
                (abs(r_tnds - 1) <= 0.1 | abs(r_bods - 1) <= 0.1)]
if (nrow(decided)) {
  decided[, verdict := ifelse(abs(r_tnds - 1) <= 0.1, "TNDS", "BODS GTFS")]
  cat("Cases the document settles:\n\n")
  for (i in seq_len(nrow(decided))) {
    d <- decided[i]
    age <- if (is.na(d$months_old)) "edition not stated" else
      sprintf("edition %s, %d months before the window",
              format(d$edition, "%d %b %Y"), d$months_old)
    cat(sprintf(paste0("* `%s` — the document implies %s journeys; TNDS %s ",
                       "(%.2f), BODS GTFS %s (%.2f). **%s is right** (%s).\n"),
                d$key, format(round(d$expected), big.mark = ","),
                format(round(d$tnds), big.mark = ","), d$r_tnds,
                format(round(d$bods_gtfs), big.mark = ","), d$r_bods,
                d$verdict, age))
  }
  cat("\n")
}
```

A ratio near 1.00 is strong evidence and a ratio far from it is a lead, not a
verdict: the document may print a different edition from the one the snapshot
carries, and `Reliable = FALSE` rows are excluded from the prose above precisely
because their published figure cannot be trusted. `Stop matched = FALSE` means
the reference stop was not found in one of the feeds, so the count is of every
same-numbered route of a matching operator and will be too high.

### The Bristol 21: the case that set the deduplication defaults

This subsection describes the state of the DfT feed **as published**. Every
figure in the tables above is counted after `UK2GTFS::gtfs_deduplicate()` has
run, so what follows is the problem that stage exists to solve, not a defect in
the numbers reported here.

*Measured directly on the DfT feed named at the top of this report, over the
first window, on 2026-07-30. The figures are quoted rather than computed here
because this report is built from the route-level validation data and not from
the feed itself; `lsoa_disagreement.md` computes the national duplication
figure as part of its own pipeline, so that is the copy that stays current.*

First Bristol's 21 is `route_id` 3700 in the DfT's GTFS. As published, over the
first window, that route_id carried **5,816 trip-days**. Reducing every trip to
the set of (stop, departure time) pairs it makes, and counting distinct
(journey, date) pairs rather than trips, leaves **3,296** — the other 2,520 are
the same bus described twice on the same day. The document implies 3,296 and
TNDS counts 3,296. The same pattern, with the same numbers, appears on
`route_id` 5320 (service 31), which points to one duplicated dataset rather than
a per-route accident.

The 21 is also the route that settled two of `gtfs_deduplicate()`'s defaults.
With `match_block = TRUE` it stayed at 5,816, because the DfT fills `block_id`
with a hash generated per dataset revision and no two copies of a journey ever
agree on it; ignoring `block_id` lands the 21 on 3,296 exactly, and First
Bristol's A1 on 6,916 against a published 6,916. `match_operator = "name"`
settled the 279, whose copies sit under two agency records for one company.
Together they took the national removal rate on this feed from 3.2% to 5.2%.

Duplication is not confined to those. Measured across the whole 2026-07-26 DfT
feed over this window, before deduplication, the same-journey-same-day rate is
several per cent of counted bus runs, concentrated rather than spread: several
`route_id`s are duplicated in their entirety. `lsoa_disagreement.md` carries the
current figure.

A caution about how that is measured, because the obvious version of the test is
wrong. Comparing trips by their (stop, time) signature alone, without regard to
date, returns 41.5% of runs — more than ten times the real figure. GTFS models a
school-term journey and its holiday twin as two trips with identical times and
*complementary* calendars, which is correct modelling and not duplication. Only
trips that run the same journey on the same **date** are counted.

```{r stopmatch}
sm <- dcast(v$feeds[window == names(v$window)[1]], key ~ source,
            value.var = "stop_matched")
setnames(sm, "key", "Route")
knitr::kable(sm, caption = "Stop matched: was the reference stop found?")
```

```{r ids}
ids <- dcast(v$feeds[window == names(v$window)[1]], key ~ source,
             value.var = "route_ids", fill = "")
setnames(ids, "key", "Route")
knitr::kable(ids, caption = "route_id(s) matched in each source")
```

Where a source lists several `route_id`s for one route they are summed: a
source that splits one service across several ids must still be compared as
one service. Several ids is in itself worth noting, since it is how duplicate
publication shows up.

Several ids is not *by itself* evidence of duplication, though, and the four
routes where TNDS reads well above BODS GTFS divide cleanly on that point. For
each of them every trip was reduced to the sorted set of its (stop, departure
time) pairs, and those signatures compared across the route's TNDS `route_id`s
(measured on `tnds_20260726_merged.zip`):

| Route | TNDS ÷ GTFS | TNDS `route_id`s | Trip patterns shared between ids | Reading |
|:------|------------:|-----------------:|:---------------------------------|:--------|
| TBALL | 2.00 | 2 | **148 of 148** | the same timetable published twice |
| TBX38 | 3.00 | 6 | **363 of 688**, as three identical pairs | three registrations, each published twice |
| SC125 | 2.00 | 2 | **0 of 731** | duplication the exact test cannot see |
| CDF62 | 3.86 | 3 | **0 of 147** | not duplication |

For TBALL and TBX38 the TNDS excess is therefore an artefact of duplicate
registration: the exact ratios of 2.00 and 3.00 follow directly from publishing
one timetable twice and one three times over, so **TNDS is overcounting them**.

The SC125 row needs correcting, and the correction is the reason
`near_duplicate_journeys.md` exists. "0 of 731" is what the exact test found and
it is accurate, but the conclusion drawn from it — that the 125 is not
duplicated — was wrong. Pair the two `route_id`s on their **stop sequence**
instead of on stops-and-times, and 230 pairs of trips run the same stops on the
same day within one minute of each other at every call, 358 within two minutes.
The service is registered twice by two Stagecoach subsidiaries, `SCCU`
(Stagecoach North West) and `SCMY` (Stagecoach Cumbria and Lancashire), and the
two registrations were written from different working timetables, so not one
journey matches exactly. TBX38 has the same thing on top of its exact
duplication: pairing its six `route_id`s the same way finds 254 pairs at zero
difference and a further 238 within a minute, so the exact test catches part of
its duplication and leaves the rest. See `near_duplicate_journeys.md` for how
common this is nationally and why the tolerance that would catch it cannot
safely be made the default.

That is as far as the trip-pattern evidence goes, and it is worth being precise
about what it does *not* establish. Compared with the documents in the section
below, TBALL's doubled TNDS figure lands nearer the document (1.12) than BODS
GTFS does (0.56), and TBX38 exceeds it in both sources. Two things make that row
unsafe to adjudicate rather than a contradiction: the reference stop is not found
in either feed for these routes, so both counts are of every same-numbered route
of a matching operator, and both documents are Word extracts whose readers are
the least trustworthy in the set (TBALL's Monday–Friday column includes
Fridays-only journeys, so its implied total is itself an overestimate). The
duplication in TNDS is established; that BODS GTFS is therefore *correct* for
these two is not.

CDF62 is the one of the four whose `route_id`s really do carry distinct trips,
and the documents settle it outright — TNDS reads 1.02 of the Cardiff 62's
published timetable and BODS GTFS 0.26, so the missing service is in BODS GTFS,
which is what England-only statutory coverage predicts for a Cardiff Bus
service. SC125 is settled the other way and by the same evidence: TNDS reads
2.28 of the document against BODS GTFS's 1.14, exactly what a service counted
twice looks like, and the near-duplicate pairing above supplies the mechanism.

## The zones where TNDS and BODS GTFS disagree most

*This section is not recomputed when the report is knitted: it is written by
`scripts/zone_pdf_validation/` from the documents and from
`data/lsoa_disagreement_2026.Rds`, and its figures are in
`data/zone_pdf_validation.csv`.*

> **Every number below this line predates the October 2026 conversion fixes.**
> It was measured over a 28-day window, on the TransXChange 2.1 edition of the
> July 2026 snapshot, and with the UK2GTFS overlap rule still discarding
> sibling files. All three of those changed; the rest of this report, which is
> knitted, reflects the change and this section does not. The 201 checks have
> been re-measured on the current pipeline @EM@
> `reports/zone_pdf_validation_refresh.md`, from
> `scripts/zone_check_refresh.R` @EM@ and that is where to read what the fixes
> did to them. This section is kept because the documents, the collection
> method, the per-cause reasoning and the reasons a timetable could not be
> obtained are all still current, and because the investigation in
> `reports/tnds_conversion_investigation.md` is written against these figures.

`lsoa_disagreement.md` ranks the 60 zones where the choice between TNDS and the
DfT's BODS GTFS changes the answer most, and tries to say from the two feeds
alone which of them is wrong. It could only do that for half of them, and one of
its largest verdicts was later reversed. This section brings the operators' own
timetables to those zones instead: for the routes that carry each zone's
disagreement, how many journeys does the published timetable say pass through
the zone in the same counting window, and which feed is nearer that number?
(28 days, as this section was measured; 14 since October 2026.)

### What was collected

121 documents covering the routes in 45 of the 60 zones, all now in
`data/example_timetables/`. They were collected for the routes that carry most
of each zone's difference: the search started from every route whose two
counts differ by 1,000 trip-runs or more in one of the zones, and for the
operators carrying many of them most of their routes into those zones were
collected.

|Operator|Source|Documents|Editions|
|:--|:--|--:|:--|
|Bluestar|operator PDFs, passenger-line-assets archive|2|in force|
|Brighton & Hove|operator PDFs, passenger-line-assets archive|7|in force|
|East Yorkshire|operator PDFs, passenger-line-assets archive|9|in force|
|First Bus|firstbus.co.uk timetable PDFs|28|in force, later edition|
|Lothian|lothianbuses.com timing sheets|10|in force|
|Metrobus|operator PDFs, passenger-line-assets archive (scanned; read by OCR)|8|in force, in force to 7 Aug|
|National Express West Midlands|nxbus.co.uk timetable PDFs (generated on download)|26|in force, later edition|
|Reading Buses|operator PDFs, passenger-line-assets archive|19|in force|
|Redline|operator PDFs, passenger-line-assets archive|3|in force|
|Stagecoach|operator PDF (already in the folder)|1|in force|
|TfL contract|TfL running schedules (already in the folder)|8|TfL schedule|

Where the operator keeps old editions, the edition in force for the window
(27 July – 23 August 2026) was used: Reading Buses, East Yorkshire, Brighton &
Hove, Bluestar, Metrobus and Redline all publish through a shared asset store
that keeps every revision, and Lothian dates its timing sheets in the file
name. National Express West Midlands and First publish only the current
edition, so most of their documents are the **autumn editions (30 August – 27
September 2026)**, a month or two after the window; the four NXWM routes
already in the folder from August (6, 14, 50, 74) are the summer edition and
are used for those. A later edition can differ from the window's by a few
journeys an hour; it cannot account for a factor of two, which is the size of
most of what follows.

Some documents could not be had. Stagecoach and Arriva sit behind a browser
check that the collection environment cannot pass, and TfL's own site refuses
it, so Chester, Leven and the TfL routes without a schedule already in the
folder (Wandsworth, Harrow, Southall, the City, Hackney) are not checked here;
trentbarton's PDFs for Derby and Burton have no text layer and were not read;
Darlington was not attempted.

### How the check works

* **The unit is the zone, not the route.** For each route the document is read
  at a timing point inside the zone — or, where the zone has none (Edinburgh's
  Salisbury Place and Drylaw, Southampton's Atherley Road), the adjacent one on
  the same corridor — counting every journey in either direction that calls
  there. That is the same quantity `lsoa_disagreement.md` counts: a vehicle
  journey touching the zone, once, however many of its stops it calls at.
* **Day types to window totals**: the window holds 16 Mondays–Thursdays, 4
  Fridays, 4 Saturdays and 4 Sundays and no English bank holiday, and it lies in
  the English and Welsh school holidays, so school-day-only journeys (First's
  `SCH`/`SD` columns) are dropped and TfL's non-schoolday schedules used.
* **Abbreviated periods are expanded** — "then every 10 minutes until",
  "at these minutes past each hour" — and a reading where an abbreviation could
  not be expanded is marked unreliable and left out of every count below
  (18 rows). NXWM's "every 10 minutes **or less**" and Reading's "**up to**
  every 9 mins" are expanded at the printed headway, so those counts are a
  floor and a ceiling respectively.
* **One check per zone and route.** Where a route's TNDS and BODS GTFS counts
  in a second zone match its counts in the checked zone to within 2%, the same
  journeys pass both, so the check is carried over (42 of the 183 checks,
  marked † below).
* **A source is right when it is within 15% of the document.** Beyond that,
  the nearer of the two on a ratio scale (a reading of 0.5 is as far off as
  2.0) is recorded as closer, not right.

### What the documents say

Across 183 zone-and-route checks in 45 zones, the document matches **TNDS in
78** (including routes BODS GTFS does not carry) and **BODS GTFS in
49** (including routes TNDS does not carry); in 56
neither source is within 15%, TNDS being the nearer in 20 and BODS
GTFS in 36.

That headline hides the real result, which is that **the answer is decided by
place, almost without exception**: within an area one source is right on
nearly every route and the other wrong by the same factor on nearly every
route, because each mechanism is an operator's or a region's publishing
pattern rather than a property of individual timetables.

|Area|Zones|Checks|TNDS ÷ doc (median)|BODS GTFS ÷ doc (median)|TNDS right|BODS GTFS right|Neither|Editions|
|:--|--:|--:|--:|--:|--:|--:|--:|:--|
|Birmingham city centre|6|42|0.94|1.67|38|0|4|in force, later edition|
|Black Country|5|13|1.00|1.81|10|0|3|in force, later edition|
|Chelmsford|6|28|0.20|2.09|0|0|28|later edition|
|Weymouth|1|5|0.28|1.83|0|0|5|later edition|
|Reading|2|19|0.24|0.98|0|15|4|in force|
|Crawley and Gatwick|3|9|0.00|1.07|0|6|3|in force, in force to 7 Aug|
|Hull|1|9|0.00|1.00|0|8|1|in force|
|Portsmouth|2|10|1.00|1.42|10|0|0|later edition|
|Edinburgh|2|12|1.00|0.50|10|0|2|in force|
|North-east London|7|15|2.00|1.00|0|15|0|TfL schedule|
|Kingston, Sunbury and Epsom|3|5|1.00|3.33|5|0|0|TfL schedule|
|Preston and Chorley|2|2|2.49|1.24|0|0|2|in force|
|Brighton|2|7|0.71|0.87|1|4|2|in force|
|Aylesbury|1|3|0.87|0.00|2|0|1|in force|
|Southampton|1|2|2.17|1.29|0|1|1|in force|
|Swansea|1|2|0.96|0.00|2|0|0|in force|

**Birmingham and the Black Country: TNDS is right, and BODS GTFS counts most
weekday buses twice.** TNDS is within 15% of the document on 48 of the
55 National Express West Midlands checks; BODS GTFS reads between
1.44 and 2.15 of it on nine in ten of them. The summer-edition
routes other than the 50, whose documents were in force for the window, show
it most cleanly:
TNDS 0.99–1.10 of the document, BODS GTFS 1.64–2.21.
This is the overlapping Monday–Friday and Monday–Thursday/Friday calendars that
`lsoa_disagreement.md` measured on route 74, now confirmed against the
operator's own timetables in 55 checks; and NXWM's own autumn timetables are printed as
separate Monday–Thursday and Friday tables, the split that BODS GTFS turns into
a second copy of the week. Route 50 is the one exception: both sources read
about twice and three times the document at Moor Street, and why is not
established.

**Chelmsford and Weymouth: neither source is right, in opposite directions.**
TNDS reads 0.10–0.52 of the First Essex and First Wessex documents
(median 0.22) — its registrations stop after six days of the
twenty-eight — and BODS GTFS reads 1.28–5.48 (median 2.09).
The BODS GTFS excess is larger than the 26–34% of duplicated journeys
`lsoa_disagreement.md` traced to `gtfs_deduplicate()` keeping copies that differ
only in arrival time at the first stop, so something else is inflating these
zones as well. These documents are September editions; the size of the
factors, not the exact ratios, is the finding. These are the zones that report
left "unresolved" because both feeds are wrong, and the documents agree with it.

**Reading: BODS GTFS is right, and TNDS counts one week of four.** BODS GTFS is
within 15% on 15 of the 19 Reading Buses checks (median
0.98), with documents in force for the window; TNDS reads a
median 0.24, the registrations ending on 2 August. This is the
one "TNDS calendars cut short" verdict, confirmed.

**Hull: BODS GTFS is right, TNDS is missing East Yorkshire.** TNDS carries none
of the 9 East Yorkshire routes checked at Hull Interchange except the
X46; BODS GTFS matches the documents to within 5% on
8 of 9, and exactly on several. The zone was "unresolved".

**Crawley and Gatwick: TNDS is missing the town network, and BODS GTFS is the
source to believe.** TNDS counts nothing for 9 of the
9 Metrobus route-and-zone checks; BODS GTFS reads a median
1.07 of the documents — 1.00–1.08 on routes 3, 10, 200 and 400,
and 1.3–1.9 on the 2 and the 100, where the OCR reading is the likelier to
have dropped journeys. The documents are scans with no
text layer and were read by OCR, so these are the least precise counts in the
section, but an absence is not a matter of precision: the service exists, and
`lsoa_disagreement.md`'s "journeys missing from TNDS" stands.

**Portsmouth: TNDS is right, and BODS GTFS overcounts.** TNDS matches First's
documents to the journey on 9 of 10 checks (ratio 0.98–1.00);
BODS GTFS reads 1.32–2.55. The zone verdict "journeys
missing from TNDS" is the wrong way round.

**Edinburgh: TNDS is right, and BODS GTFS carries half the service.** TNDS is
within 15% of Lothian's timing sheets on 10 of 12 checks, exactly on
three; BODS GTFS reads 0.40–0.60, half or a little under on every
route. "Journeys missing from BODS GTFS" stands, and it is a uniform halving
rather than missing routes.

**North-east London: BODS GTFS is right, and TNDS counts every bus twice.**
Against TfL's running schedules BODS GTFS reads 0.99–1.01 and TNDS
1.99–2.01 on routes 20, 167, 215, 275 and 462. This is the exact-2.00
group of `lsoa_disagreement.md` — duplicate TNDS registrations — and the zones it
labelled "journeys missing from BODS GTFS" (Hainault Grove, Audleigh Place,
Loughton Station) are the wrong way round.

**Kingston, Sunbury and Epsom: TNDS is right, and BODS GTFS reads two to four
times the schedules.** TNDS 1.00–1.05, BODS GTFS 2.00–3.99
on routes 216, 235 and 406 — the second London pattern, in the opposite
direction from the first. "Journeys missing from TNDS" is overturned.

**Preston and Chorley: neither, BODS GTFS nearer.** The 125 reads TNDS
2.41–2.57 and BODS GTFS 1.20–1.28 of the document at both
bus stations — the duplicate Stagecoach registration in TNDS found earlier in
this report, now seen at zone level.

**The rest are mixed, and mostly about coverage.** In Brighton, BODS GTFS lacks
routes 50, 12A and 18, which TNDS carries (50 to the journey), while TNDS lacks
the 20, which BODS GTFS carries to within 2%; on the shared routes BODS GTFS is
nearer. In Aylesbury, BODS GTFS carries none of Redline's 4, 130 and 300 and
TNDS reads 0.84–0.89 of them. In Swansea TNDS matches First Cymru's 4 and X6
and BODS GTFS has neither, as Welsh services need not appear in it. In
Southampton, read at Central Station for want of a timing point in the zone,
BODS GTFS matches Bluestar's 17 and reads the 18 half again high, while TNDS
reads 1.8–2.5 of both.

### What that means for the zone verdicts

Weighting each checked route by how much of its zone's difference it carries,
the documents find **TNDS right in 19 zones and BODS GTFS right in
14**; in the other 12
neither source is right, BODS GTFS being the nearer in 10
(Chelmsford, Weymouth, Preston, Chorley, Southampton) and TNDS in
2. By size, the zones where TNDS is right or nearer hold
52% of the checked zones' combined difference and
those where BODS GTFS is hold 48%.

Set against the verdicts in `lsoa_disagreement.md`:

* of the zones it blamed on **TNDS**, the documents agree in
  5 and disagree in 10 —
  every disagreement being a zone where BODS GTFS publishes more journeys than
  run (Birmingham, Portsmouth, Sunbury Cross, Epsom);
* of those it blamed on **BODS GTFS**, they agree in 3
  and disagree in 5 — the disagreements being
  TNDS's own duplicate registrations (north-east London, Chorley, Southampton);
* of the **unresolved** zones checked, the documents settle
  7 for TNDS (the Black Country bus stations, Kingston,
  Swansea) and 6 for BODS GTFS (Hull, Walthamstow
  and Woodford, Brighton's North Street), and confirm that in
  9 more
  both sources are wrong at once — which is what "unresolved" was meant to say.

So the arithmetic in that report — journeys against operating days — reliably
finds *where* the feeds disagree and *what kind* of disagreement it is, but
"journeys missing from" one source is as often a duplicate in the other. The
caution already in that report — that a flat daily profile with a
journey-count gap should send the reader to the registrations before deciding
which source is short — is borne out: ten of the fifteen checked zones it
blamed on TNDS are the other way round.

|Zone|Locality|Difference|Verdict in lsoa_disagreement.md|Routes checked|Checked routes' difference ÷ zone's|Document sides with|
|:--|:--|--:|:--|--:|--:|:--|
|E01033620|Church Centre|-74,064|journeys missing from TNDS|18|73%|TNDS|
|E01034091|Bus Station|-55,922|unresolved|13|83%|BODS GTFS (closer)|
|E01033617|Albert Street|-50,692|journeys missing from TNDS|5|30%|TNDS|
|E01034092|Cathedral|-48,614|unresolved|5|23%|BODS GTFS (closer)|
|E01033415|Friar Street|-45,946|TNDS calendars cut short|16|69%|BODS GTFS|
|E01033140|Parkway|-43,567|unresolved|3|12%|BODS GTFS (closer)|
|E01033561|Moor St Selfridges|-40,808|journeys missing from TNDS|7|60%|TNDS|
|E01033615|Markets|-34,416|journeys missing from TNDS|4|20%|TNDS|
|E01034313|Wolverhampton Bus Station|-32,595|unresolved|5|40%|TNDS|
|E01010102|West Bromwich Bus Station|-28,356|unresolved|2|47%|TNDS|
|E01002968|Cromwell Road Bus Station|-27,482|unresolved|2|50%|TNDS|
|E01033420|Kings Road|-25,762|TNDS calendars cut short|3|16%|BODS GTFS|
|E01017032|City Shops South|-24,570|journeys missing from TNDS|6|74%|TNDS|
|E01010125|Chelmsley Interchange|-24,168|journeys missing from TNDS|6|79%|TNDS|
|E01031575|North Terminal Bus Station|-24,160|journeys missing from TNDS|3|54%|BODS GTFS|
|E01020554|Kings Statue|-23,405|unresolved|5|81%|BODS GTFS (closer)|
|E01031585|Bus Station|-23,280|journeys missing from TNDS|5|80%|BODS GTFS|
|E01021592|Cockney Corner|-23,014|unresolved|3|67%|BODS GTFS (closer)|
|E01017034|The Hard Interchange|-22,595|journeys missing from TNDS|4|70%|TNDS|
|E01031583|Manor Royal Central|-21,972|journeys missing from TNDS|1|33%|BODS GTFS|
|E01010368|Walsall Bus Station|-21,947|unresolved|2|25%|TNDS|
|E01009757|Evolve Campus|-21,136|unresolved|3|30%|TNDS|
|E01010106|New Street|-20,992|unresolved|1|28%|TNDS|
|E01021587|Skerry Rise|-20,987|unresolved|2|45%|BODS GTFS (closer)|
|E01030751|Sunbury Cross|-20,924|journeys missing from TNDS|2|69%|TNDS|
|E01034947|Barford Street|-20,600|journeys missing from TNDS|2|44%|TNDS (closer)|
|E01034290|High Street|-19,745|journeys missing from TNDS|1|30%|TNDS|
|E01033104|Hull Interchange|-19,312|unresolved|9|83%|BODS GTFS|
|E01021542|Hospital|-18,421|unresolved|2|51%|BODS GTFS (closer)|
|W01001955|Orchard Street|5,672|unresolved|2|64%|TNDS|
|E01021767|Hainault Grove|5,916|journeys missing from BODS GTFS|2|100%|BODS GTFS|
|E01017191|Atherley Road|6,011|journeys missing from BODS GTFS|2|65%|BODS GTFS (closer)|
|S01014958|Drylaw Police Station|6,568|journeys missing from BODS GTFS|4|98%|TNDS|
|E01021766|Audleigh Place|6,592|journeys missing from BODS GTFS|2|100%|BODS GTFS|
|E01016952|Churchill Square|7,124|unresolved|1|47%|TNDS (closer)|
|E01032955|Aylesbury Bus Stn|7,188|journeys missing from BODS GTFS|3|60%|TNDS|
|E01016969|North Street|7,476|unresolved|6|174%|BODS GTFS|
|E01004377|Chelmsford Rd /Woodford New Rd|7,532|unresolved|2|97%|BODS GTFS|
|E01024940|Bus Station|7,540|journeys missing from BODS GTFS|1|58%|BODS GTFS (closer)|
|E01003670|St Aubyns School|7,540|unresolved|2|97%|BODS GTFS|
|E01003750|Chingford Lane (IG8)|7,590|unresolved|2|96%|BODS GTFS|
|E01021782|Loughton Station|7,920|journeys missing from BODS GTFS|2|77%|BODS GTFS|
|E01033223|Bus Station|8,001|unresolved|1|50%|BODS GTFS (closer)|
|E01004397|Walthamstow Bus Station|10,248|unresolved|3|99%|BODS GTFS|
|S01014636|Salisbury Place|13,004|journeys missing from BODS GTFS|8|96%|TNDS|

### Every check

`Edition`: *in force* is the edition valid for the window; *later edition*
started after it; *TfL schedule* is a TfL running schedule current in July
2026. † marks a check carried over from another zone with the same journeys.
The same table, with the per-day counts and the document file names, is in
`data/zone_pdf_validation.csv`.

|Zone|Locality|Route|Edition|Document|TNDS|BODS GTFS|TNDS ÷ doc|BODS ÷ doc|Verdict|
|:--|:--|:--|:--|--:|--:|--:|--:|--:|:--|
|E01033617|Albert Street|14|in force|4,952|5,440|10,096|1.10|2.04|TNDS right †|
|E01033617|Albert Street|94|later edition|4,612|4,224|7,892|0.92|1.71|TNDS right|
|E01033617|Albert Street|95|later edition|4,660|4,160|7,848|0.89|1.68|TNDS right †|
|E01033617|Albert Street|X13|later edition|1,984|1,748|3,108|0.88|1.57|TNDS right|
|E01033617|Albert Street|X51|later edition|4,344|4,104|6,064|0.94|1.40|TNDS right|
|E01034947|Barford Street|35|later edition|3,872|3,688|5,808|0.95|1.50|TNDS right †|
|E01034947|Barford Street|50|in force|6,008|11,488|18,460|1.91|3.07|neither; TNDS closer †|
|E01010125|Chelmsley Interchange|14|in force|4,952|5,440|10,096|1.10|2.04|TNDS right|
|E01010125|Chelmsley Interchange|94|later edition|4,612|4,224|7,892|0.92|1.71|TNDS right †|
|E01010125|Chelmsley Interchange|95|later edition|4,660|4,160|7,848|0.89|1.68|TNDS right|
|E01010125|Chelmsley Interchange|96|later edition|1,756|1,756|3,096|1.00|1.76|TNDS right|
|E01010125|Chelmsley Interchange|97|later edition|5,284|4,776|8,992|0.90|1.70|TNDS right †|
|E01010125|Chelmsley Interchange|X12|later edition|2,228|2,104|3,716|0.94|1.67|TNDS right|
|E01033620|Church Centre|101|later edition|4,032|3,920|6,468|0.97|1.60|TNDS right (unreliable reading)|
|E01033620|Church Centre|126|later edition|2,760|2,760|4,616|1.00|1.67|TNDS right|
|E01033620|Church Centre|14|in force|4,952|5,440|10,096|1.10|2.04|TNDS right †|
|E01033620|Church Centre|23|later edition|4,800|4,324|6,788|0.90|1.41|TNDS right|
|E01033620|Church Centre|24|later edition|4,748|4,240|6,992|0.89|1.47|TNDS right|
|E01033620|Church Centre|4|later edition|4,728|4,628|7,288|0.98|1.54|TNDS right|
|E01033620|Church Centre|4A|later edition|2,592|2,672|4,432|1.03|1.71|TNDS right|
|E01033620|Church Centre|5|later edition|1,712|1,712|2,844|1.00|1.66|TNDS right|
|E01033620|Church Centre|6|in force|4,956|5,256|10,952|1.06|2.21|TNDS right|
|E01033620|Church Centre|74|in force|8,272|8,224|13,592|0.99|1.64|TNDS right|
|E01033620|Church Centre|82|later edition|4,260|3,976|6,532|0.93|1.53|TNDS right|
|E01033620|Church Centre|87|later edition|4,548|4,272|7,576|0.94|1.67|TNDS right|
|E01033620|Church Centre|9|later edition|4,952|4,336|8,276|0.88|1.67|TNDS right|
|E01033620|Church Centre|94|later edition|4,612|4,224|7,892|0.92|1.71|TNDS right †|
|E01033620|Church Centre|95|later edition|4,660|4,160|7,848|0.89|1.68|TNDS right †|
|E01033620|Church Centre|97|later edition|5,284|4,776|8,992|0.90|1.70|TNDS right|
|E01033620|Church Centre|X20|later edition|1,936|1,528|2,724|0.79|1.41|neither; TNDS closer|
|E01033620|Church Centre|X21|later edition|3,092|2,692|4,480|0.87|1.45|TNDS right|
|E01033620|Church Centre|X22|later edition|2,960|2,608|4,312|0.88|1.46|TNDS right|
|E01033620|Church Centre|X8|later edition|1,744|2,208|3,868|1.27|2.22|neither; TNDS closer (unreliable reading)|
|E01033615|Markets|35|later edition|3,872|3,688|5,808|0.95|1.50|TNDS right|
|E01033615|Markets|X20|later edition|1,936|1,528|2,724|0.79|1.41|neither; TNDS closer †|
|E01033615|Markets|X21|later edition|3,092|2,692|4,480|0.87|1.45|TNDS right †|
|E01033615|Markets|X22|later edition|2,960|2,608|4,312|0.88|1.46|TNDS right †|
|E01033561|Moor St Selfridges|14|in force|4,976|5,440|10,096|1.09|2.03|TNDS right|
|E01033561|Moor St Selfridges|17|later edition|2,900|2,780|5,008|0.96|1.73|TNDS right|
|E01033561|Moor St Selfridges|17A|later edition|1,956|1,904|3,236|0.97|1.65|TNDS right|
|E01033561|Moor St Selfridges|35|later edition|3,872|3,688|5,808|0.95|1.50|TNDS right †|
|E01033561|Moor St Selfridges|50|in force|6,008|11,488|18,460|1.91|3.07|neither; TNDS closer|
|E01033561|Moor St Selfridges|94|later edition|4,612|4,224|7,892|0.92|1.71|TNDS right †|
|E01033561|Moor St Selfridges|95|later edition|4,660|4,160|7,848|0.89|1.68|TNDS right|
|E01009757|Evolve Campus|126|later edition|2,780|2,780|4,636|1.00|1.67|TNDS right|
|E01009757|Evolve Campus|82|later edition|1,268|1,620|2,660|1.28|2.10|neither; TNDS closer|
|E01009757|Evolve Campus|87|later edition|4,484|4,316|7,656|0.96|1.71|TNDS right|
|E01010106|New Street|79|later edition|4,628|4,262|10,062|0.92|2.17|TNDS right †|
|E01010368|Walsall Bus Station|529|later edition|5,040|5,580|9,136|1.11|1.81|TNDS right|
|E01010368|Walsall Bus Station|X51|later edition|4,480|4,196|6,156|0.94|1.37|TNDS right|
|E01010102|West Bromwich Bus Station|74|in force|10,748|10,900|18,540|1.01|1.72|TNDS right|
|E01010102|West Bromwich Bus Station|79|later edition|4,628|4,262|10,062|0.92|2.17|TNDS right|
|E01034313|Wolverhampton Bus Station|16|later edition|2,440|2,204|3,504|0.90|1.44|TNDS right|
|E01034313|Wolverhampton Bus Station|529|later edition|5,040|5,580|9,184|1.11|1.82|TNDS right|
|E01034313|Wolverhampton Bus Station|59|later edition|3,488|4,584|5,616|1.31|1.61|neither; TNDS closer|
|E01034313|Wolverhampton Bus Station|79|later edition|4,676|4,330|10,274|0.93|2.20|TNDS right|
|E01034313|Wolverhampton Bus Station|9|later edition|1,452|1,892|3,080|1.30|2.12|neither; TNDS closer|
|E01034091|Bus Station|170|later edition|1,468|291|2,109|0.20|1.44|neither; BODS GTFS closer|
|E01034091|Bus Station|333|later edition|1,056|165|1,350|0.16|1.28|neither; BODS GTFS closer|
|E01034091|Bus Station|336|later edition|1,000|140|1,441|0.14|1.44|neither; BODS GTFS closer|
|E01034091|Bus Station|351|later edition|844|200|2,324|0.24|2.75|neither; BODS GTFS closer|
|E01034091|Bus Station|C1|later edition|2,324|817|8,575|0.35|3.69|neither; TNDS closer|
|E01034091|Bus Station|C10|later edition|1,564|364|4,740|0.23|3.03|neither; BODS GTFS closer|
|E01034091|Bus Station|C2|later edition|1,404|617|6,675|0.44|4.75|neither; TNDS closer|
|E01034091|Bus Station|C3|later edition|720|348|3,948|0.48|5.48|neither; TNDS closer|
|E01034091|Bus Station|C5|later edition|1,112|517|4,843|0.46|4.36|neither; TNDS closer|
|E01034091|Bus Station|C7|later edition|960|171|2,532|0.18|2.64|neither; BODS GTFS closer|
|E01034091|Bus Station|C8|later edition|1,176|276|3,444|0.23|2.93|neither; BODS GTFS closer|
|E01034091|Bus Station|X10|later edition|1,280|279|2,341|0.22|1.83|neither; BODS GTFS closer|
|E01034091|Bus Station|X30|later edition|2,340|231|6,764|0.10|2.89|neither; BODS GTFS closer|
|E01034092|Cathedral|333|later edition|1,056|165|1,350|0.16|1.28|neither; BODS GTFS closer †|
|E01034092|Cathedral|336|later edition|1,000|140|1,441|0.14|1.44|neither; BODS GTFS closer †|
|E01034092|Cathedral|351|later edition|844|200|2,324|0.24|2.75|neither; BODS GTFS closer †|
|E01034092|Cathedral|700|later edition|1,612|403|2,961|0.25|1.84|neither; BODS GTFS closer †|
|E01034092|Cathedral|C9|later edition|1,212|286|4,290|0.24|3.54|neither; BODS GTFS closer|
|E01021592|Cockney Corner|170|later edition|1,468|291|2,109|0.20|1.44|neither; BODS GTFS closer †|
|E01021592|Cockney Corner|C1|later edition|4,044|805|8,435|0.20|2.09|neither; BODS GTFS closer †|
|E01021592|Cockney Corner|C2|later edition|1,404|617|6,671|0.44|4.75|neither; TNDS closer †|
|E01021542|Hospital|170|later edition|1,468|291|2,109|0.20|1.44|neither; BODS GTFS closer †|
|E01021542|Hospital|C1|later edition|4,044|805|8,435|0.20|2.09|neither; BODS GTFS closer|
|E01033140|Parkway|333|later edition|1,056|165|1,350|0.16|1.28|neither; BODS GTFS closer †|
|E01033140|Parkway|336|later edition|1,000|140|1,441|0.14|1.44|neither; BODS GTFS closer †|
|E01033140|Parkway|700|later edition|1,612|403|2,961|0.25|1.84|neither; BODS GTFS closer|
|E01021587|Skerry Rise|170|later edition|1,468|291|2,109|0.20|1.44|neither; BODS GTFS closer †|
|E01021587|Skerry Rise|C1|later edition|4,044|805|8,435|0.20|2.09|neither; BODS GTFS closer|
|E01020554|Kings Statue|1|later edition|3,496|756|5,663|0.22|1.62|neither; BODS GTFS closer|
|E01020554|Kings Statue|10|later edition|3,012|665|4,790|0.22|1.59|neither; BODS GTFS closer|
|E01020554|Kings Statue|2|later edition|3,320|918|6,079|0.28|1.83|neither; BODS GTFS closer|
|E01020554|Kings Statue|4|later edition|1,128|450|3,253|0.40|2.88|neither; TNDS closer|
|E01020554|Kings Statue|8|later edition|660|342|2,322|0.52|3.52|neither; TNDS closer|
|E01033415|Friar Street|1|in force|1,648|412|1,648|0.25|1.00|BODS GTFS right|
|E01033415|Friar Street|11|in force|2,456|614|2,456|0.25|1.00|BODS GTFS right|
|E01033415|Friar Street|16|in force|2,452|603|2,412|0.25|0.98|BODS GTFS right|
|E01033415|Friar Street|17|in force|6,840|1,629|6,516|0.24|0.95|BODS GTFS right|
|E01033415|Friar Street|18|in force|1,644|401|1,604|0.24|0.98|BODS GTFS right|
|E01033415|Friar Street|21|in force|3,688|761|3,044|0.21|0.83|neither; BODS GTFS closer|
|E01033415|Friar Street|25|in force|1,368|363|1,452|0.27|1.06|BODS GTFS right|
|E01033415|Friar Street|26|in force|3,916|950|3,800|0.24|0.97|BODS GTFS right|
|E01033415|Friar Street|28|in force|1,920|369|1,476|0.19|0.77|neither; BODS GTFS closer|
|E01033415|Friar Street|29|in force|1,428|374|1,496|0.26|1.05|BODS GTFS right|
|E01033415|Friar Street|3|in force|2,728|759|3,036|0.28|1.11|BODS GTFS right|
|E01033415|Friar Street|33|in force|2,492|623|2,492|0.25|1.00|BODS GTFS right|
|E01033415|Friar Street|4|in force|2,212|410|1,640|0.19|0.74|neither; BODS GTFS closer|
|E01033415|Friar Street|5|in force|4,412|1,102|4,408|0.25|1.00|BODS GTFS right|
|E01033415|Friar Street|50|in force|996|359|1,436|0.36|1.44|neither; BODS GTFS closer (unreliable reading)|
|E01033415|Friar Street|500|in force|892|607|2,428|0.68|2.72|neither; TNDS closer (unreliable reading)|
|E01033415|Friar Street|6|in force|4,860|1,052|4,208|0.22|0.87|BODS GTFS right|
|E01033415|Friar Street|600|in force|1,304|745|2,980|0.57|2.29|neither; TNDS closer (unreliable reading)|
|E01033415|Friar Street|9|in force|940|207|828|0.22|0.88|BODS GTFS right †|
|E01033420|Kings Road|3|in force|2,728|759|3,036|0.28|1.11|BODS GTFS right †|
|E01033420|Kings Road|4|in force|2,212|410|1,640|0.19|0.74|neither; BODS GTFS closer †|
|E01033420|Kings Road|9|in force|940|207|828|0.22|0.88|BODS GTFS right|
|E01031585|Bus Station|10|in force|7,208|0|7,172|0.00|1.00|TNDS absent|
|E01031585|Bus Station|100|in force|2,664|0|3,580|0.00|1.34|TNDS absent; BODS GTFS off|
|E01031585|Bus Station|2|in force|1,908|0|3,620|0.00|1.90|TNDS absent; BODS GTFS off|
|E01031585|Bus Station|20|in force|0|0|1,568|—|—|neither; BODS GTFS closer (unreliable reading)|
|E01031585|Bus Station|3|in force|2,672|0|2,768|0.00|1.04|TNDS absent|
|E01031585|Bus Station|4|in force|4,360|0|1,664|0.00|0.38|TNDS absent; BODS GTFS off (unreliable reading)|
|E01031585|Bus Station|400|in force to 7 Aug|1,356|0|1,456|0.00|1.07|TNDS absent|
|E01031585|Bus Station|5|in force|4,360|0|1,664|0.00|0.38|TNDS absent; BODS GTFS off (unreliable reading)|
|E01031583|Manor Royal Central|10|in force|7,200|0|7,172|0.00|1.00|TNDS absent|
|E01031583|Manor Royal Central|100|in force|0|0|3,556|—|—|neither; BODS GTFS closer (unreliable reading)|
|E01031583|Manor Royal Central|20|in force|0|0|3,140|—|—|neither; BODS GTFS closer (unreliable reading)|
|E01031583|Manor Royal Central|200|in force|0|0|2,320|—|—|neither; BODS GTFS closer (unreliable reading)|
|E01031583|Manor Royal Central|3|in force|0|0|2,776|—|—|neither; BODS GTFS closer (unreliable reading)|
|E01031583|Manor Royal Central|400|in force to 7 Aug|0|0|1,408|—|—|neither; BODS GTFS closer (unreliable reading)|
|E01031575|North Terminal Bus Station|10|in force|6,816|0|7,172|0.00|1.05|TNDS absent|
|E01031575|North Terminal Bus Station|100|in force|2,632|0|3,588|0.00|1.36|TNDS absent; BODS GTFS off|
|E01031575|North Terminal Bus Station|20|in force|0|0|3,140|—|—|neither; BODS GTFS closer (unreliable reading)|
|E01031575|North Terminal Bus Station|200|in force|2,148|0|2,320|0.00|1.08|TNDS absent|
|E01031575|North Terminal Bus Station|3|in force|0|0|2,776|—|—|neither; BODS GTFS closer (unreliable reading)|
|E01031575|North Terminal Bus Station|4|in force|0|0|1,664|—|—|neither; BODS GTFS closer (unreliable reading)|
|E01031575|North Terminal Bus Station|400|in force to 7 Aug|0|0|1,600|—|—|neither; BODS GTFS closer (unreliable reading)|
|E01031575|North Terminal Bus Station|5|in force|0|0|1,664|—|—|neither; BODS GTFS closer (unreliable reading)|
|E01033104|Hull Interchange|104|in force|1,568|0|1,568|0.00|1.00|TNDS absent|
|E01033104|Hull Interchange|35|in force|1,528|0|1,528|0.00|1.00|TNDS absent|
|E01033104|Hull Interchange|45|in force|1,952|0|2,032|0.00|1.04|TNDS absent|
|E01033104|Hull Interchange|51|in force|1,924|0|1,924|0.00|1.00|TNDS absent|
|E01033104|Hull Interchange|54|in force|1,544|0|1,544|0.00|1.00|TNDS absent|
|E01033104|Hull Interchange|56|in force|1,968|0|1,964|0.00|1.00|TNDS absent|
|E01033104|Hull Interchange|57|in force|2,144|0|2,144|0.00|1.00|TNDS absent|
|E01033104|Hull Interchange|58|in force|1,724|0|1,764|0.00|1.02|TNDS absent|
|E01033104|Hull Interchange|X46|in force|1,240|834|2,456|0.67|1.98|neither; TNDS closer|
|E01017032|City Shops South|1|later edition|4,120|4,120|10,474|1.00|2.54|TNDS right|
|E01017032|City Shops South|2|later edition|4,556|4,556|6,103|1.00|1.34|TNDS right|
|E01017032|City Shops South|3|later edition|4,188|4,148|10,666|0.99|2.55|TNDS right|
|E01017032|City Shops South|7|later edition|3,336|3,336|4,394|1.00|1.32|TNDS right|
|E01017032|City Shops South|8|later edition|3,484|3,484|4,939|1.00|1.42|TNDS right|
|E01017032|City Shops South|X3|later edition|1,316|1,296|0|0.98|0.00|BODS GTFS absent|
|E01017034|The Hard Interchange|1|later edition|4,120|4,120|10,474|1.00|2.54|TNDS right †|
|E01017034|The Hard Interchange|2|later edition|4,556|4,556|6,103|1.00|1.34|TNDS right †|
|E01017034|The Hard Interchange|3|later edition|4,188|4,148|10,666|0.99|2.55|TNDS right †|
|E01017034|The Hard Interchange|8|later edition|3,484|3,484|4,935|1.00|1.42|TNDS right †|
|S01014958|Drylaw Police Station|21|in force|3,320|3,640|1,848|1.10|0.56|TNDS right|
|S01014958|Drylaw Police Station|27|in force|3,300|3,444|1,756|1.04|0.53|TNDS right|
|S01014958|Drylaw Police Station|29|in force|2,184|2,600|1,316|1.19|0.60|neither; TNDS closer|
|S01014958|Drylaw Police Station|37|in force|3,616|3,476|1,784|0.96|0.49|TNDS right|
|S01014636|Salisbury Place|29|in force|2,808|2,728|1,344|0.97|0.48|TNDS right|
|S01014636|Salisbury Place|3|in force|4,212|4,212|2,100|1.00|0.50|TNDS right|
|S01014636|Salisbury Place|31|in force|4,096|4,096|2,048|1.00|0.50|TNDS right|
|S01014636|Salisbury Place|37|in force|3,164|3,504|1,792|1.11|0.57|TNDS right|
|S01014636|Salisbury Place|47|in force|2,848|2,280|1,144|0.80|0.40|neither; TNDS closer|
|S01014636|Salisbury Place|49|in force|2,660|2,660|1,284|1.00|0.48|TNDS right|
|S01014636|Salisbury Place|7|in force|3,104|2,940|1,464|0.95|0.47|TNDS right|
|S01014636|Salisbury Place|8|in force|2,292|2,488|1,220|1.09|0.53|TNDS right|
|E01021766|Audleigh Place|167|TfL schedule|2,712|5,424|2,712|2.00|1.00|BODS GTFS right †|
|E01021766|Audleigh Place|275|TfL schedule|3,920|7,840|3,960|2.00|1.01|BODS GTFS right †|
|E01004377|Chelmsford Rd /Woodford New Rd|20|TfL schedule|3,392|6,804|3,392|2.01|1.00|BODS GTFS right †|
|E01004377|Chelmsford Rd /Woodford New Rd|275|TfL schedule|3,920|7,840|3,960|2.00|1.01|BODS GTFS right †|
|E01003750|Chingford Lane (IG8)|20|TfL schedule|3,392|6,804|3,392|2.01|1.00|BODS GTFS right †|
|E01003750|Chingford Lane (IG8)|275|TfL schedule|3,920|7,840|3,960|2.00|1.01|BODS GTFS right †|
|E01021767|Hainault Grove|167|TfL schedule|2,712|5,424|2,712|2.00|1.00|BODS GTFS right †|
|E01021767|Hainault Grove|462|TfL schedule|3,204|6,408|3,204|2.00|1.00|BODS GTFS right|
|E01021782|Loughton Station|167|TfL schedule|2,712|5,424|2,712|2.00|1.00|BODS GTFS right|
|E01021782|Loughton Station|20|TfL schedule|3,392|6,804|3,392|2.01|1.00|BODS GTFS right|
|E01003670|St Aubyns School|20|TfL schedule|3,392|6,804|3,392|2.01|1.00|BODS GTFS right †|
|E01003670|St Aubyns School|275|TfL schedule|3,920|7,840|3,960|2.00|1.01|BODS GTFS right †|
|E01004397|Walthamstow Bus Station|20|TfL schedule|3,392|6,804|3,392|2.01|1.00|BODS GTFS right|
|E01004397|Walthamstow Bus Station|215|TfL schedule|2,836|5,640|2,820|1.99|0.99|BODS GTFS right|
|E01004397|Walthamstow Bus Station|275|TfL schedule|3,920|7,840|3,960|2.00|1.01|BODS GTFS right|
|E01002968|Cromwell Road Bus Station|216|TfL schedule|2,616|2,616|10,432|1.00|3.99|TNDS right|
|E01002968|Cromwell Road Bus Station|406|TfL schedule|2,632|2,756|8,752|1.05|3.33|TNDS right|
|E01034290|High Street|406|TfL schedule|2,632|2,756|8,752|1.05|3.33|TNDS right|
|E01030751|Sunbury Cross|216|TfL schedule|2,616|2,616|10,432|1.00|3.99|TNDS right|
|E01030751|Sunbury Cross|235|TfL schedule|6,640|6,640|13,280|1.00|2.00|TNDS right|
|E01024940|Bus Station|125|in force|3,664|8,816|4,408|2.41|1.20|neither; BODS GTFS closer|
|E01033223|Bus Station|125|in force|3,116|8,008|4,004|2.57|1.28|neither; BODS GTFS closer|
|E01016952|Churchill Square|18|in force|1,548|3,320|0|2.14|0.00|BODS GTFS absent; TNDS off|
|E01016969|North Street|12|in force|4,304|3,072|4,190|0.71|0.97|BODS GTFS right|
|E01016969|North Street|12A|in force|3,720|2,512|0|0.68|0.00|BODS GTFS absent; TNDS off|
|E01016969|North Street|14|in force|3,460|2,112|3,349|0.61|0.97|BODS GTFS right|
|E01016969|North Street|20|in force|3,264|0|3,199|0.00|0.98|TNDS absent|
|E01016969|North Street|25|in force|5,196|6,392|4,508|1.23|0.87|BODS GTFS right|
|E01016969|North Street|50|in force|3,028|3,028|0|1.00|0.00|BODS GTFS absent|
|E01032955|Aylesbury Bus Stn|130|in force|1,704|1,520|0|0.89|0.00|BODS GTFS absent|
|E01032955|Aylesbury Bus Stn|300|in force|1,584|1,380|0|0.87|0.00|BODS GTFS absent|
|E01032955|Aylesbury Bus Stn|4|in force|1,676|1,400|0|0.84|0.00|BODS GTFS absent; TNDS off|
|E01017191|Atherley Road|17|in force|2,600|4,684|2,789|1.80|1.07|BODS GTFS right|
|E01017191|Atherley Road|18|in force|1,948|4,940|2,950|2.54|1.51|neither; BODS GTFS closer|
|W01001955|Orchard Street|4|in force|2,440|2,256|0|0.92|0.00|BODS GTFS absent|
|W01001955|Orchard Street|X6|in force|1,360|1,360|0|1.00|0.00|BODS GTFS absent|

### Limits

* A later edition is not the window's timetable. For NXWM and First the ratios
  are evidence of factors of two and four, not of exact agreement; where an
  in-force edition exists (NXWM 6, 14, 74) it agrees with the later one to
  within a few per cent.
* The reference timing point stands for the zone. Where the zone has no timing
  point of its own, the adjacent one on the same road is used, and a journey
  that turns off between them would be miscounted.
* Readings are mechanical, with expansion of abbreviated periods. The reader
  was checked against the route-level counts earlier in this report (NXWM 14
  and 6: within two journeys a day) and by hand on a few documents (First's
  C1, Reading's 5 and 17); readings it could not complete are excluded rather
  than estimated. The Metrobus documents are scans read by OCR.
* The zones not checked are mostly those whose operators could not be reached
  (Stagecoach, Arriva) and the London routes without a TfL schedule in the
  folder — `missing_timetables.md` lists every document still wanted, with
  links; the scripts that collected and read everything here
  are in `scripts/zone_pdf_validation/`.

## Notes on each document

Figures quoted *inside* these notes are **selection rationale**, not
measurements: they come from the February 2026 comparison window that these
documents were picked from. That window has since been dropped from the
comparison (its TNDS side is distorted by registrations expiring inside it), and
in any case it is not the window counted here. Where a note disagrees with the
Window totals table above, **the table is the measurement** — the Wigan `AN320`
is noted as TNDS reading 1.96× BODS GTFS, but on this July snapshot the two
sources agree exactly.

```{r notes}
n <- rbindlist(lapply(v$routes, function(r)
  data.table(Route = r$key,
             Operator = r$operator,
             Skipped = isTRUE(r$skip),
             Note = r$note)))
knitr::kable(n)
```
